---
output: github_document
---

# tsn

`tsn` constructs networks directly from time-series geometry. Its core
constructor handles whole-series distances, sliding-window distances, natural
visibility, horizontal visibility, and visibility networks over discretized
states. Focused verbs expose discretization, trend classification, and
Nestimate transition-network models. Core network results are list-backed
objects containing a tidy dyad table and carrying the
`netobject`/`cograph_network` classes for downstream use.

Core construction uses base R only. `Nestimate` is an optional dependency for
the `ts_tna()` transition-network family. Network plots use `cograph`
exclusively; the source-series diagnostic plot remains available without it.

> Not to be confused with `tsnet` (Bayesian graphical VAR), `tsna` (temporal
> social networks), or `ts2net` (a related but distinct time-series-to-network
> toolkit). `tsn` emphasizes a compact, dependency-light R interface with
> irregular-time visibility support, fifteen discretizers, and optional
> transition-network integration.

## Installation

The package is not currently published on CRAN. Install the development release
from GitHub:

```r
install.packages("remotes")
remotes::install_github("mohsaqr/tsn")
```

## One-liner API

The fastest path is a bare series and a method string, or the `vg()` verb:

```r
library(tsn)

tsn(c(3, 1, 4, 2, 5, 3, 6, 2, 7), "hvg")        # horizontal visibility graph
tsn(c(3, 1, 4, 2, 5, 3, 6, 2, 7), "nvg")        # natural visibility graph
tsn(c(3, 1, 4, 2, 5, 3, 6, 2, 7), "ordinal")    # ordinal (permutation) network
tsn(c(3, 1, 4, 2, 5, 3, 6, 2, 7), "distance")   # distance network

vg(c(3, 1, 4, 2, 5, 3, 6, 2, 7))                # natural visibility graph
vg(c(3, 1, 4, 2, 5, 3, 6, 2, 7), "horizontal")  # horizontal visibility graph
```

`vg()` is the discoverable verb for visibility graphs; it forwards every
visibility-relevant argument to `tsn()` and returns an equivalent object (with
the wrapper call retained as provenance).
The method-string shortcuts (`"hvg"`, `"nvg"`, and every discretizer name)
keep working.

## Method map

| Goal | Main call | Choices |
|---|---|---|
| Compare complete series or windows | `tsn(method = "distance")` | 15 distances; full, nearest, threshold, percentile, or Gaussian connections |
| Map time-series geometry | `vg()` or `tsn(method = "visibility")` | natural/horizontal, directed, penetrable, elapsed-time limit/decay |
| Collapse visibility to states | `tsn(unit = "state")` | 15 discretizers; count, sum, mean, or maximum aggregation |
| Inspect state assignments | `discretize()` | scalar and group-safe temporal methods |
| Classify rolling direction | `trend()` | OLS, Theil-Sen, Spearman, Kendall, or growth factor |
| Model state transitions | `ts_tna()` family | probability, frequency, co-occurrence, or attention networks via Nestimate |

## The full verb

```r
series <- list(
  first = c(1, 2, 3, 2, 1),
  second = c(1, 1, 2, 3, 5),
  third = c(5, 4, 3, 2, 1)
)

tsn(
  data = series,
  method = "distance",
  unit = "series",
  distance = "dtw",
  connect = "nearest",
  neighbors = 2
)
```

Long data uses explicit, readable column names:

```r
observations <- data.frame(
  student = rep(c("A", "B", "C"), each = 6),
  week = rep(seq_len(6), times = 3),
  score = c(1:6, 2:7, 7:2)
)

tsn(
  data = observations,
  value = "score",
  id = "student",
  time = "week",
  method = "distance",
  unit = "series",
  distance = "correlation",
  connect = "full"
)
```

Network options include `chain = TRUE` (connect only consecutive series or
windows, a transition chain), `directed = TRUE`, and `normalize = TRUE`.

## Uniform results and plotting

Every result supports the same standard methods and two plot views:

```r
network <- tsn(c(3, 1, 4, 2, 5, 3, 6, 2, 7), "hvg")

network
summary(network)
as.data.frame(network)               # the tidy dyad table
as.data.frame(network, what = "series")
as.matrix(network)                   # the weighted adjacency matrix

plot(network)                        # cograph network view
plot(network, layout = "spring")     # any cograph layout/option
plot(network, "series")              # tsn source-series view
```

For distance networks, weights are similarities: larger values indicate closer
units. Visibility weights are edge strengths; state aggregation by `"count"` or
`"sum"` can therefore exceed one.

Network rendering belongs exclusively to `cograph`; a `tsn` result is already
a `cograph_network`, and `plot(network)` delegates to `cograph::splot()`:

```r
install.packages("cograph")
cograph::splot(network)
```

## Trends and discretization

```r
walk <- cumsum(rnorm(120))
trend(walk, window = 15, slope = "ols")           # rolling trend classification

discretize(walk, method = "quantile", n_states = 3)
discretize(walk, method = "ordinal", m = 3)
discretize(walk, method = "width", transform = "log",
           labels = c("low", "mid", "high"))
```

`discretize()` exposes all fifteen state methods (`threshold`, `width`,
`quantile`, `kde`, `kmeans`, `gaussian`, `hclust`, `ordinal`, `symbolic`,
`change_points`, `entropy`, `magnitude`, `adaptive_magnitude`,
`percentile_magnitude`, `dtw`); `tsn(unit = "state")` routes through the same
engine, so the two agree exactly.

Temporal discretizers are group-safe: ordinal patterns, adaptive-magnitude
features, and DTW windows are computed separately within each selected series,
then mapped to one shared state vocabulary.

State visibility plots support both time-oriented and value-oriented context.
The package dataset can be used directly; `series` selects a participant inside
`tsn()`:

```{r eval=FALSE}
data(steps)
state_network <- tsn(
  data = steps,
  value = "steps",
  id = "id",
  time = "day",
  series = 536,
  unit = "state",
  discretization = "quantile"
)

plot(state_network, "series", overlay = "vertical")
plot(state_network, "series", overlay = "horizontal")
```

`"vertical"` shades contiguous state runs along time. `"horizontal"` shades
state value bands across the y-axis. Both use base graphics.

Mixed wide data is equally direct:

```{r eval=FALSE}
data(motivation)
tsn(
  motivation,
  series = c("autonomy", "competence", "relatedness", "mood"),
  method = "distance",
  distance = "correlation"
)
```

## State transition networks

The transition family discretizes each series, preserves a common state
alphabet, and delegates network estimation to Nestimate:

```{r eval=FALSE}
install.packages("Nestimate")

walks <- list(
  first = cumsum(rnorm(80)),
  second = cumsum(rnorm(80))
)
model <- ts_tna(walks, n_states = 3,
                labels = c("low", "mid", "high"))
model$weights

per_series <- series_networks(model)
plot(model, ribbon = TRUE)                    # series + per-series networks
plot(model, network = "summary")              # pooled transition network
Nestimate::net_centrality(model)
```

Inferential methods retain their usual sampling requirements. In particular,
sequence bootstrap is not informative for a model built from only one
sequence.

See `vignette("pleasure-all-functions")` for a tutorial that applies every
exported tsn function to the packaged `motivation$pleasure` series.

## Package boundaries

`tsn` owns time-series-to-network construction and the bridge from numeric
series to state sequences. Nestimate owns transition-network estimation and
inference; tsn calls its builders rather than reimplementing them. cograph owns
generic graph conversion, analytics, layouts, and network rendering.
