---
output: github_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  fig.align = "center",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 150,
  out.width = "100%"
)

library(tsn)
set.seed(2026)
```

# tsn

`tsn` turns one or many time series into a network. What differs between
analyses is only **what the nodes are**:

```
                    ┌─► states ──────────────────► transition network
one or many  ───────┤
time series         └─► series, windows, or ─────► geometry network
                        observations
```

- **Nodes are states.** `discretize()` maps the numeric values onto a small
  state alphabet, and the network records how the series moves between those
  states over time. This is a transition network.
- **Nodes are the measurements themselves** — whole series, sliding windows,
  or individual observations. Distance or visibility between them supplies the
  edges directly, with no discretization involved.

Nine exported verbs cover both:

| What you want | Verb | Choices |
|---|---|---|
| Model state transitions | `ts_tna()`, `ts_ftna()`, `ts_cna()`, `ts_atna()` | probability, frequency, co-occurrence, or attention networks via Nestimate |
| Split a pooled model | `series_networks()` | one network per series |
| Discretize into states | `discretize()` (or `tsn(unit = "state")`) | 15 discretizers, group-safe across series |
| Classify rolling direction | `trend()` | Theil–Sen (default), OLS, Spearman, or Kendall slope; growth factor |
| Build a geometry network | `tsn()`, `vg()` | 15 distances; natural or horizontal visibility; full, nearest, threshold, percentile, or Gaussian connection |

Core construction uses base R only. `Nestimate` is an optional dependency for
the `ts_tna()` transition-network family; network plots delegate to `cograph`
(with `igraph`), while the source-series view needs neither.

> Not to be confused with `tsnet` (Bayesian graphical VAR), `tsna` (temporal
> social networks), or `ts2net` (a related but distinct time-series-to-network
> toolkit). `tsn` emphasizes a compact, dependency-light R interface with
> irregular-time visibility support, fifteen discretizers, and optional
> transition-network integration.

## Installation

The package is not currently published on CRAN. Install the development
release from GitHub:

```r
install.packages("remotes")
remotes::install_github("mohsaqr/tsn")
```

## Transition networks between states

The `ts_tna()` family goes from raw measurements to a transition network in
one call: it discretizes each series, keeps one shared state alphabet across
them, and delegates the estimation to `Nestimate` (a Suggests dependency).
`ts_tna()` fits transition probabilities, `ts_ftna()` raw frequencies,
`ts_cna()` co-occurrences, and `ts_atna()` attention-weighted transitions.

The packaged `steps` data records daily step counts. Participant 35 has
missing days, so we model the complete observations for two participants:

```{r transitions, message = FALSE}
data(steps)

complete <- subset(steps, !is.na(steps))

model <- ts_tna(
  complete,
  value = "steps",
  id = "id",
  time = "day",
  series = c(536, 88),
  n_states = 3,
  labels = c("low", "moderate", "high")
)

model
```

Days are sticky: in the pooled model a `low` day is followed by another `low`
day with probability 0.578, and a `high` day by another `high` day with
probability 0.605. The pooled state alphabet is learned from both series
together, so the two participants are placed on one common scale;
`series_networks()` then splits the model into per-series networks:

```{r per-series}
series_networks(model)
```

The returned collection supports `print()`, `summary()`, `as.data.frame()`,
and `plot()` directly (`plot()` takes `series = "<name>"` when the collection
holds more than one model). The combined plot stacks each participant's
shaded series next to their own transition network:

```{r transition-plot, fig.width = 9, fig.height = 5.5, fig.alt = "Two rows, one per participant: each shows the daily step series with a state ribbon strip underneath and that participant's three-state transition network beside it, with edge weights printed on the arrows."}
plot(model, ribbon = TRUE, overlay = "none")
```

Because the result is a Nestimate model, its downstream tools apply
unchanged (for example `Nestimate::net_centrality(model)`). Inferential
methods retain their usual sampling requirements; in particular, sequence
bootstrap is not informative for a model built from a single sequence.

## From values to states: `discretize()` and `trend()`

`discretize()` is the bridge from numeric series to state sequences. The
packaged `steps` data records daily step counts; participant 536 contributes
265 complete days.

```{r discretize}
states <- discretize(
  steps,
  value = "steps",
  id = "id",
  time = "day",
  series = 536,
  method = "quantile",
  n_states = 3,
  labels = c("low", "moderate", "high")
)

summary(states)
```

Fifteen discretizers are available (`threshold`, `width`, `quantile`, `kde`,
`kmeans`, `gaussian`, `hclust`, `ordinal`, `symbolic`, `change_points`,
`entropy`, `magnitude`, `adaptive_magnitude`, `percentile_magnitude`,
`dtw`), and the temporal ones are group-safe: ordinal patterns,
adaptive-magnitude features, and DTW windows are computed separately within
each series, then mapped to one shared state vocabulary.

`trend()` is a companion discretizer for direction: it classifies each
observation by its rolling slope.

```{r trend}
movement <- trend(
  steps,
  value = "steps",
  id = "id",
  time = "day",
  series = 536,
  window = 14
)

summary(movement)
```

## Visibility networks: observations as nodes

A visibility graph maps a single series onto a network of its observations:
two time points are connected when a straight sightline over the intervening
values joins them. The packaged `motivation` data holds 4,871 experience
samples; the first 60 pleasure ratings are enough to see the idea.

```{r visibility}
data(motivation)

pleasure <- head(motivation, 60)
network <- vg(pleasure, series = "pleasure")

summary(network)
```

The 60 observations become 60 nodes joined by 143 sightlines. Every network
has two plot views: `plot(x)` renders the network itself through `cograph`,
and `plot(x, "series")` shows the source series it was built from.

```{r visibility-network, fig.alt = "Natural visibility graph of sixty pleasure ratings, drawn as a spring-layout network with nodes sized by degree; a few high-degree hubs correspond to peak observations that see far along the series."}
plot(network)
```

```{r visibility-series, fig.height = 3, fig.alt = "The sixty pleasure ratings the visibility graph was built from, plotted as a line over observation order."}
plot(network, "series")
```

`vg(x)` builds a natural visibility graph and `vg(x, "horizontal")` a
horizontal one; the `tsn()` method strings `"nvg"` and `"hvg"` are sugar for
the same calls. Visibility options include `directed`, `penetrable`
(sightlines may cross a few points), `limit` (a maximum elapsed time), and
`decay`; when a `time` column is supplied, sightlines and elapsed-time rules
use the observed — possibly irregular — spacing rather than row positions.

## State networks from visibility

`tsn(unit = "state")` runs the same discretization engine and then collapses
the visibility graph onto the states: nodes are states, and each edge
aggregates (by default, sums) the sightlines between occurrences of a state
pair.

```{r state-network}
state_network <- tsn(
  steps,
  value = "steps",
  id = "id",
  time = "day",
  series = 536,
  unit = "state",
  discretization = "quantile",
  n_states = 3
)

summary(state_network)
```

The series view shades the states over the raw values, either as horizontal
value bands (`overlay = "horizontal"`) or as vertical runs along time:

```{r state-overlay, fig.height = 3.2, fig.alt = "Daily step counts for participant 536 with horizontal shading marking the three quantile state bands across the y-axis."}
plot(state_network, "series", overlay = "horizontal")
```

## Distance networks: series and windows as nodes

With `method = "distance"`, the nodes are whole series (or sliding windows)
and edge weights are similarities: larger means closer. Comparing five
motivation variables as complete series takes one call:

```{r distance-series}
affect <- tsn(
  motivation,
  series = c("pleasure", "autonomy", "competence", "relatedness", "mood"),
  method = "distance",
  distance = "correlation"
)

summary(affect)
```

```{r distance-network, fig.height = 5, fig.alt = "Correlation-distance network of five motivation variables, drawn with labeled nodes; all ten pairs are connected and autonomy and competence form the strongest tie."}
plot(affect, labels = TRUE)
```

All ten pairs are connected because `connect = "full"` is the default;
autonomy–competence is the closest pair. Fifteen distances are available
(`euclidean`, `manhattan`, `maximum`, `canberra`, `minkowski`, `binary`,
`cosine`, `correlation`, `spearman`, `dtw`, `ccf`, `nmi`, `voi`,
`event_sync`, `van_rossum`), and `connect` sparsifies the result by nearest
neighbors, a threshold, a percentile, or a Gaussian kernel.

The same verb compares sliding windows of one series, which turns a single
trajectory into a network of its own epochs:

```{r distance-windows}
windows <- tsn(
  pleasure,
  series = "pleasure",
  method = "distance",
  distance = "dtw",
  window = 12,
  step = 4,
  connect = "nearest",
  neighbors = 2
)

summary(windows)
```

The 60 ratings yield 13 overlapping windows, and keeping each window's two
nearest neighbors retains 17 of the 78 possible dyads. The default `window`
is 10% of the series length, so choose `step` relative to how long the series
is. Further options include `chain = TRUE` (connect only consecutive windows
or series), `directed = TRUE`, and `normalize = TRUE`.

## One object, one grammar

However it was built, a `tsn` network is the same kind of object: a
list-backed result carrying the `netobject`/`cograph_network` classes, with a
tidy dyad table inside. The standard verbs work everywhere:

```r
summary(network)                         # one tidy row describing the network
as.data.frame(network)                   # the dyad table
as.data.frame(network, what = "series")  # the source observations
as.matrix(network)                       # the weighted adjacency matrix

plot(network)                            # network view, rendered by cograph
plot(network, "series")                  # source-series view, base graphics
```

`plot()` on a network applies readable defaults (spring layout, degree-scaled
nodes); any `cograph` argument passes straight through and overrides them,
for example `plot(network, layout = "circle", labels = TRUE)`.

## Vignettes

- `vignette("pleasure-all-functions")` applies every exported tsn function to
  the packaged `motivation` pleasure series, end to end.
- `vignette("plotting-time-series-networks")` covers the plotting surface:
  the network and source-series views, state overlays, and transition-model
  plots.

## Package boundaries

`tsn` owns time-series-to-network construction and the bridge from numeric
series to state sequences. `Nestimate` owns transition-network estimation and
inference; tsn calls its builders rather than reimplementing them. `cograph`
owns generic graph conversion, analytics, layouts, and network rendering: a
`tsn` result is already a `cograph_network`, and `plot(network)` delegates to
`cograph::splot()`.
