---
title: "Grouped Transition Models: One Network per Condition"
author: "tsn package authors"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Grouped Transition Models: One Network per Condition}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 110,
  out.width = "100%"
)
library(tsn)
has_nestimate <- requireNamespace("Nestimate", quietly = TRUE)
has_cograph <- requireNamespace("cograph", quietly = TRUE) &&
  requireNamespace("igraph", quietly = TRUE)
```

# The question

Does the same person move between states differently under different
conditions? The `motivation` study measured momentary pleasure up to seven
times a day for almost three years, and each measurement records where it was
taken: at home, at work, in personal time, or elsewhere. A single transition
network estimated from all of it would answer how pleasure moves between
`low`, `mid`, and `high` *on average* — and would erase exactly the
distinction the context column exists to draw.

The `group` argument of `ts_tna()` and its siblings answers the conditional
question instead: one network per context, all cut from one shared state
alphabet, packaged so that Nestimate's grouped verbs compare them directly.

```{r libs, eval=has_nestimate}
library(tsn)
library(Nestimate)

data(motivation)
```

# One call, one network per context

`group` names a column of the data. Every level of that column becomes its
own fully estimated `ts_tna` model.

```{r grouped-full, eval=has_nestimate}
by_context <- ts_tna(
  motivation,
  series = "pleasure",
  group = "task_context_type",
  labels = c("low", "mid", "high")
)

by_context
```

Four contexts, four networks, and one of them is already suspicious: `Other`
carries a single edge with weight 1.000. `as.data.frame()` with
`what = "groups"` is the tidy index of how much data stands behind each
network, and it explains that edge.

```{r grouped-index, eval=has_nestimate}
as.data.frame(by_context, what = "groups")
```

`Other` holds three observations in two sequences — one transition in total.
A transition probability estimated from one transition is not a result; it is
a level of the column that happens to exist. Every grouped analysis should
start by reading this table. Here the fix is to leave `Other` out and rebuild.

```{r grouped, eval=has_nestimate}
contexts <- subset(motivation, task_context_type != "Other")

by_context <- ts_tna(
  contexts,
  series = "pleasure",
  group = "task_context_type",
  labels = c("low", "mid", "high")
)

as.data.frame(by_context, what = "groups")
```

Three contexts remain, each backed by 1,309 to 2,235 observations. The
`sequences` column deserves attention before anything else is read, and the
next two sections explain the two design rules that produce it.

# Why not subset the data and build each model by hand?

Subsetting looks equivalent: keep the `Home` rows, call `ts_tna()` on them,
repeat per context. It is not equivalent, and it fails in two distinct ways.

```{r home-alone, eval=has_nestimate}
home_alone <- ts_tna(
  subset(motivation, task_context_type == "Home"),
  series = "pleasure",
  labels = c("low", "mid", "high")
)

summary(series_networks(home_alone))
```

## First failure: the subset welds broken time back together

The hand-built model holds all 1,324 home measurements as **one** sequence —
the single row above. But the study alternates contexts from one beep to the
next, so the home measurements are not consecutive: the grouped index showed
them forming **832** separate runs. One unbroken sequence of 1,324
observations contains 1,323 transitions; 832 runs of 1,324 observations
contain only 1,324 − 832 = 492. The hand-built model therefore counts 831
transitions — sixty-three percent of its total — between measurements that
were never adjacent in time. Each one joins the end of one home visit to the
start of the next, across an intervening trip to work or elsewhere, and each
one is dynamics that never happened.

The grouped model never counts them. A sequence is cut at every change of
group, so a transition is only ever counted between genuinely consecutive
measurements within one context. The transition *across* a context change is
dropped from both groups, because it belongs to neither.

## Second failure: each subset learns its own thresholds

The second failure is quieter. The default `discretization = "quantile"`
cuts at the tertiles of whatever data it is given, so the hand-built home
model learns cut points from home measurements alone:

```{r home-alone-dist, eval=has_nestimate}
state_distribution(home_alone)
```

By construction, roughly a third of home measurements land in each state —
37.3% `low`, 33.3% `high`. Now the same measurements inside the grouped
model, which cut the states from the **pooled** series before splitting:

```{r grouped-dist, eval=has_nestimate}
state_distribution(by_context)
```

On the shared scale, home is not one-third `low`: it is **62.5%** `low` and
only **6.6%** `high`. The hand-built model erased that fact by defining
`low` as "low for a home measurement". The grouped model keeps one
definition of `low` for everyone, which is the only definition under which
"Home has more low pleasure than Personal" is even a well-formed claim.

The comparison across the three contexts is now worth reading on its own:
pleasure at home is `low` on 62.5% of measurements and `high` on 6.6%;
personal time reverses the ordering (`high` 45.6%, `low` 19.3%); work sits
between the two on all three states. The mosaic view draws that table:

```{r mosaic, eval=has_nestimate, fig.width=7, fig.height=4, fig.alt="Mosaic plot of state proportions by context: the low-pleasure band dominates the Home column, the high-pleasure band dominates the Personal column, and Work sits between them."}
mosaic_plot(by_context)
```

The same discipline is available without `group` by passing explicit
`breaks`, and that is the right tool when thresholds carry outside meaning
(the companion vignette
`vignette("nestimate-workflow", package = "tsn")` cuts step counts at 5,000
and 10,000 steps for exactly that reason). `group` automates the discipline
when the comparison, not the threshold, is the point.

# The collection is data

Every view of the grouped model is a data frame with a leading `group`
column. The default view is the edge list:

```{r edges, eval=has_nestimate}
head(as.data.frame(by_context), 9)
```

Because all groups share one alphabet, a single edge can be read across
contexts:

```{r one-edge, eval=has_nestimate}
subset(as.data.frame(by_context), from == "low" & to == "high")
```

The probability of jumping from `low` directly to `high` is 0.040 at home,
0.351 in personal time, and 0.291 at work: at home, low pleasure almost
never resolves upward in one step. `what = "series"` returns the
per-observation table behind the networks, with the run-level sequence IDs.

```{r series, eval=has_nestimate}
head(as.data.frame(by_context, what = "series"))
```

`summary()` adds Nestimate's network-metric panel, one column per group:

```{r summary, eval=has_nestimate}
summary(by_context)
```

Most rows are structural constants for a full 3-state network, but one is
not: the standard deviation of out-strength falls from 0.70 at home through
0.45 in personal time to 0.21 at work. Home's outgoing probabilities are the
most polarized — from any state, one destination dominates — while work's
are the most evenly spread. That single row anticipates the entropy analysis
below.

# Drawing the networks

There is no single picture of a grouped model; `plot()` draws one selected
group with the full `plot.ts_tna()` surface.

```{r plot-home, eval=has_nestimate && has_cograph, fig.width=6, fig.height=5, fig.alt="Transition network of the Home context: three nodes labelled low, mid, and high, with the heaviest directed edges pointing into the low state and a strong low-to-low self-loop."}
plot(by_context, group = "Home", type = "network")
```

```{r plot-personal, eval=has_nestimate && has_cograph, fig.width=6, fig.height=5, fig.alt="Transition network of the Personal context: three nodes labelled low, mid, and high, with the heaviest directed edges pointing into the high state and a strong high-to-high self-loop."}
plot(by_context, group = "Personal", type = "network")
```

```{r plot-work, eval=has_nestimate && has_cograph, fig.width=6, fig.height=5, fig.alt="Transition network of the Work context: three nodes labelled low, mid, and high, with edge weights spread more evenly than in the Home and Personal networks."}
plot(by_context, group = "Work", type = "network")
```

Because the three networks share one node set, the figures are directly
comparable: the heavy edges converge on `low` at home (`low → low` is
0.762) and on `high` in personal time (`high → high` is 0.664), while the
work network's weights are visibly more even — no edge above 0.518.

# Dynamics, not composition alone

The state distributions above describe *where* the contexts spend their
time. The networks describe *how they move*, and the two can disagree.

## Mean first passage times

`passage_time()` reports, per group, the expected number of steps to reach
each state from each other state (Kemeny & Snell, 1976).

```{r passage, eval=has_nestimate}
passage_time(by_context)
```

At home, `low` is reached from anywhere within about two steps, while
`high` takes sixteen to eighteen — low pleasure at home is an attractor.
Personal time is the mirror image: `high` is two to three steps away from
everything, `low` seven to ten. Work reaches `high` in about three steps.
The stationary distributions agree with the composition table, which they
should — but the passage times add the direction of pull, not only the
destination.

## Entropy: how predictable is each context?

```{r entropy, eval=has_nestimate}
transition_entropy(by_context)
```

The entropy rate is in bits, with `log2(3) = 1.585` as the ceiling for three
states. Home, at 1.09, is the most predictable context: knowing the current
state says a lot about the next one (mostly, that it will be `low` again).
Work, at 1.51, runs close to the ceiling — from any state, the next work
measurement is nearly a three-way coin toss. Structure and pleasantness are
different axes: home is the most *predictable* context and also the least
pleasant one.

## Centrality per group

```{r centrality, eval=has_nestimate}
context_centrality <- net_centrality(by_context)

context_centrality
```

The in-strength ordering flips between contexts: `low` dominates at home
(0.92), `high` dominates in personal time (0.79). The message printed above
the tables matters for any transition network: self-transitions are excluded
by default, because the diagonal is usually the largest entry and would
otherwise drown the between-state structure. The result has its own plot
method, which puts the three groups on one panel per measure:

```{r centrality-plot, eval=has_nestimate, fig.width=7, fig.height=4.5, fig.alt="Faceted centrality chart comparing in-strength, betweenness, and diffusion of the three states across the Home, Personal, and Work networks, showing the in-strength ordering reversing between Home and Personal."}
plot(context_centrality)
```

## Edge betweenness

`net_edge_betweenness()` scores each edge by the shortest paths that cross
it. On a grouped model it runs group-wise:

```{r betweenness, eval=has_nestimate}
net_edge_betweenness(by_context)
```

Home and Personal each concentrate their traffic on five edges while Work
spreads it across six with uniform load — the betweenness restatement of the
polarized-versus-even pattern. One caveat: this verb returns a plain
`netobject_group` (the `ts_tna_group` subclass is dropped), so apply tsn's
tidy accessors before this step, not after it.

# Are the differences real?

## Pairwise permutation tests

`permutation()` on a grouped model runs every pairwise comparison, shuffling
sequences between the two groups to build a null distribution per edge.

```{r permutation, eval=has_nestimate}
comparison <- permutation(by_context, iter = 1000, seed = 2026)

comparison
```

`summary()` is the tidy edge-level table across all three pairs:

```{r permutation-summary, eval=has_nestimate}
summary(comparison)
```

The pattern is sharp. Home differs from Personal on eight of nine edges and
from Work on eight of nine, most at the smallest p-value the permutation
resolution allows (0.001). Personal and Work are much closer: only three
edges separate them, and the largest of those differences is the retention
of `high` — `high → high` is 0.66 in personal time against 0.52 at work
(p = 0.001). The conclusion is not "everything differs from everything":
home has genuinely distinct dynamics, while personal time and work share
most of their structure except how long a good state lasts.

## A metric panel for one pair

`compare_model()` picks two groups by name and returns a panel of matrix
comparison metrics.

```{r compare, eval=has_nestimate}
compare_model(by_context, i = "Home", j = "Personal")
```

The Pearson correlation between the two weight matrices is **−0.54**: the
home and personal-time networks are not merely different, they are inversely
related — the transitions that are strong in one are weak in the other. The
mean absolute edge difference is 0.30 and the largest single gap is 0.52
(the `high → high` retention). Always name `i` and `j`: with more than two
groups there is no meaningful default pair.

## Bootstrap and Bayesian certainty per group

`bootstrap_network()` on the collection resamples sequences within each
group; `certainty()` reaches a similar verdict from a Dirichlet posterior,
without resampling.

```{r bootstrap, eval=has_nestimate}
bootstrap_network(by_context, iter = 200, seed = 2026)
```

```{r certainty, eval=has_nestimate}
set.seed(2026)
certainty(by_context)
```

Work, with 976 sequences and 2,235 observations, keeps eight of nine
bootstrap edges and all nine posterior-certain ones; Home and Personal keep
three to four under either criterion. The groups table from the start of
this vignette predicted this: equal-looking observation counts do not buy
equal precision when the observations arrive in short runs, because
resampling operates on runs, not measurements. Reading `sequences` first is
not a formality.

## Is first order enough, per group?

Each group's network assumes the next state depends only on the current one.
`markov_order_test()` (after Anderson & Goodman, 1957) checks that
assumption group by group.

```{r markov, eval=has_nestimate}
set.seed(2026)
markov_order_test(by_context)
```

Home and Personal select order 1 on BIC and the permutation LRT; Work's
permutation test flags order 2 while BIC still prefers 1. The first-order
models above are adequate summaries for home and personal time; for work,
yesterday's state carries some extra information, so its network should be
read as a one-step summary rather than the full process.

# Pruning and stability

`net_prune()` applies group-wise and returns the same class, so a pruned
collection flows back into every verb above.

```{r prune, eval=has_nestimate}
net_prune(by_context, method = "threshold", threshold = 0.3)
```

Each context keeps six of its nine edges at a 0.3 threshold. Prune after
testing, not before: the permutation table above is exactly the kind of
result a premature prune would have deleted edges from.

Two stability checks ask whether the results would survive a smaller sample.
`centrality_stability()` tracks the centrality *ordering* under case
dropping; `casedrop_reliability()` tracks the edge weights themselves (both
follow the case-drop logic of Epskamp, Borsboom & Fried, 2018).

```{r stability, eval=has_nestimate}
set.seed(2026)
centrality_stability(by_context, iter = 100)
```

```{r casedrop, eval=has_nestimate}
context_reliability <- casedrop_reliability(by_context, iter = 100,
                                            seed = 2026)

context_reliability
```

```{r casedrop-plot, eval=has_nestimate, fig.width=7, fig.height=4, fig.alt="Case-dropping stability curves for the three context networks, showing the correlation between full-sample and subsampled edge weights declining as larger fractions of sequences are dropped."}
plot(context_reliability)
```

In-strength and out-strength orderings hold at 0.6 to 0.8 across groups, and
every group's edge weights carry a CS-coefficient of 0.6 — above the 0.5 bar
for acceptable, short of the 0.7 bar for excellent. Work's betweenness
column is `NA` because its betweenness scores are degenerate (the centrality
table above showed zeros); a coefficient cannot be computed for an ordering
that does not exist. The verdict across both checks: these grouped estimates
tolerate losing roughly half their sequences, and claims resting on finer
distinctions than that should not be made from this sample.

# What to watch for

Four failure modes, all of which announce themselves:

* **A level with almost no data.** `Other` above: it gets a network because
  it is a level, and the network means nothing. The `groups` table is the
  guard; read `sequences` and `observations` before edges.
* **A group column that alternates too fast.** Every change of group cuts a
  sequence, so a column that flips at almost every observation leaves runs
  of length one and no transitions at all. In the extreme case — a group
  whose observations are *never* adjacent — the model warns that the group's
  network is empty. This study's contexts alternate heavily (about 2,630
  runs across 4,868 observations) and still leave 487 to 1,259 transitions
  per context; a faster-alternating column would not.
* **A group whose rows do not survive discretization.** Temporal
  discretizers such as `ordinal` drop tail rows that cannot fill an
  embedding window. If that removes a group entirely, the model warns that
  the group was lost to discretization. A group you excluded yourself via
  `series` is not warned about — it was your decision.
* **Comparing a grouped model with a hand-subset model.** The two are not
  estimates of the same thing, for both reasons in the section above.

# When to use which

* **`group`** — the condition already exists as a column, there may be more
  than two levels, and the comparison itself is the point. Pooled
  discretization and boundary-safe sequence cutting come built in.
* **Two models plus `permutation(x, y)`** — the split is not a column but a
  derived rule (a date cutoff, an external roster), and there are exactly
  two sides. Fix the state alphabet yourself with explicit `breaks`, or the
  comparison inherits the threshold artifact. The companion vignette
  `vignette("nestimate-workflow", package = "tsn")` runs this design on the
  step-count data.
* **`series_networks()`** — the unit of comparison is the individual series
  (one network per participant), not a condition shared across series.
* **A single pooled model** — no conditioning variable at all; the average
  dynamics are the question.

# References

Anderson, T. W., & Goodman, L. A. (1957). Statistical inference about
Markov chains. *The Annals of Mathematical Statistics*, 28(1), 89–110.

Epskamp, S., Borsboom, D., & Fried, E. I. (2018). Estimating psychological
networks and their accuracy: A tutorial paper. *Behavior Research Methods*,
50(1), 195–212.

Kemeny, J. G., & Snell, J. L. (1976). *Finite Markov Chains*. Springer.
