---
title: "The Nestimate Compatibility Map"
author:
  - name: "Mohammed Saqr"
    affiliation: "University of Eastern Finland"
  - name: "Sonsoles López-Pernas"
    affiliation: "University of Eastern Finland"
  - name: "Manuel J. Gómez"
    affiliation: "University of Murcia"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{The Nestimate Compatibility Map}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 110,
  out.width = "100%"
)
library(tsn)
has_nestimate <- requireNamespace("Nestimate", quietly = TRUE)
has_cograph <- requireNamespace("cograph", quietly = TRUE) &&
  requireNamespace("igraph", quietly = TRUE)
```

# One object, two packages

A `ts_tna()` model is not exported *to* Nestimate; it *is* a Nestimate
object. This vignette maps what that buys, verb by verb: which parts of
Nestimate's analysis surface accept a tsn transition model directly, what
question each verb answers, what each result draws, and which parts do not
apply and why. The companion vignette
`vignette("nestimate-workflow", package = "tsn")` runs a single analysis end
to end; this one is the reference you consult when deciding which verb to
reach for.

The running example is the model from that workflow: the 36 students of the
`srl` daily panel, whose `effort` reports are cut at 40 and 70 on the 0-100
scale so the states mean something outside the sample.

```{r build, eval=has_nestimate}
library(tsn)
library(Nestimate)

data(srl)
students <- subset(srl, !is.na(effort))

effort_model <- ts_tna(
  students,
  value = "effort",
  id = "name",
  time = "day",
  discretization = "threshold",
  breaks = c(40, 70),
  labels = c("low", "moderate", "high")
)

class(effort_model)
```

The class chain is the whole compatibility story. `ts_tna` is the tsn layer
(it remembers the source series and discretization), `netobject` is what
every Nestimate verb dispatches on, and `cograph_network` is what
`cograph::splot()` renders. Nestimate's own validator agrees:

```{r validate, eval=has_nestimate}
validate_netobject(effort_model)
```

And because the object is a `cograph_network`, the model itself is one
`plot()` away:

```{r plot-model, eval=has_nestimate && has_cograph, fig.width=6.5, fig.height=5, fig.alt="Directed transition network with low, moderate, and high effort nodes; edge labels show transition probabilities, with the heaviest self-loop on the high state."}
plot(effort_model, type = "network")
```

The strong diagonal is the first substantive fact: every state prefers
itself, `high` most of all at 0.575.

# Reading the model as data

Both packages provide accessors, and they answer slightly different
questions. tsn's `as.data.frame()` returns every state pair including
self-transitions; `as.matrix()` returns the row-stochastic weight matrix.

```{r tsn-accessors, eval=has_nestimate}
head(as.data.frame(effort_model))
```

Nestimate's extractors return the same estimates in its own shapes:

```{r extract, eval=has_nestimate}
extract_transition_matrix(effort_model)

extract_initial_probs(effort_model)

extract_edges(effort_model)
```

One difference matters for transition networks: `extract_edges()` **excludes
self-transitions** — it returns the six between-state edges, sorted by
weight, and omits the diagonal. For this model the diagonal is where most of
the probability lives (`high → high` is 0.575), so use `as.data.frame()` or
`extract_transition_matrix()` whenever persistence is part of the question,
and `extract_edges()` when only between-state movement is.

The initial-state probabilities say the panel starts engaged: 41.7% of
students' first recorded day is a `high`-effort day, 19.4% a `low` one.

# Where the process spends its time

`state_distribution()` and `state_frequencies()` count observations per
state — the marginal composition, before any dynamics.

```{r frequencies, eval=has_nestimate}
state_frequencies(effort_model)
```

High-effort days make up 41.7% of the panel and low-effort days 24.4%. Three
displays draw the composition from different angles. The mosaic is the
proportion bar:

```{r mosaic, eval=has_nestimate, fig.width=6.5, fig.height=4, fig.alt="Mosaic plot of the effort model's state composition, with high and moderate days forming most of the area and low days a narrower band."}
mosaic_plot(effort_model)
```

`sequence_plot()` is the sequence carpet — one row per student, one coloured
cell per day:

```{r carpet, eval=has_nestimate, fig.width=7, fig.height=5, fig.alt="Sequence heatmap with one row per student and one coloured cell per day, showing same-state spells as horizontal streaks of colour."}
sequence_plot(effort_model)
```

The streaks are persistence made visible: students hold a state for spells,
not single days. `distribution_plot()` stacks the day-by-day state mix of the
whole panel:

```{r composition-time, eval=has_nestimate, fig.width=7, fig.height=4, fig.alt="Stacked area chart of the panel's state proportions across the 156 course days, with bands for low, moderate, and high effort."}
distribution_plot(effort_model)
```

This is composition over course time — the view in which a first-to-second
half shift would appear, and the workflow vignette's period comparison shows
there is none. `plot_state_frequencies()` offers a fourth, marimekko-style
variant of the same information.

# Which states and edges carry the structure?

`net_centrality()` ranks states. Self-transitions are excluded by default —
for a transition network the diagonal would otherwise dominate every
strength measure.

```{r centrality, eval=has_nestimate}
effort_centrality <- net_centrality(effort_model)

effort_centrality
```

```{r centrality-plot, eval=has_nestimate, fig.width=7, fig.height=4, fig.alt="Centrality chart of the three states showing high and moderate with similar in-strength and low well behind."}
plot(effort_centrality)
```

`high` and `moderate` receive similar between-state traffic (in-strengths of
0.60 and 0.57) while `low` receives markedly less (0.38); the betweenness
column is uniformly zero because in a fully connected three-state network no
state is an obligatory bridge.

`net_edge_betweenness()` asks the same question of edges:

```{r betweenness, eval=has_nestimate}
effort_betweenness <- net_edge_betweenness(effort_model)

effort_betweenness
```

```{r betweenness-plot, eval=has_nestimate, fig.width=6.5, fig.height=4.5, fig.alt="Edge-betweenness network view in which every between-state edge carries the same unit load."}
plot(effort_betweenness)
```

Every between-state edge carries the same load of 1: each lies on exactly one
shortest path, and none is a privileged route. That uniformity is itself a
description — movement between effort states does not funnel through any
single corridor.

# How the process moves

## Mean first passage times

`passage_time()` converts the transition matrix into expected travel times
between states (Kemeny & Snell, 1976).

```{r passage, eval=has_nestimate}
effort_passage <- passage_time(effort_model)

effort_passage
```

```{r passage-plot, eval=has_nestimate, fig.width=6.5, fig.height=4.5, fig.alt="Heatmap of mean first passage times between the three states, with the longest times into the low state."}
plot(effort_passage)
```

A `moderate` or `high` day is reached from anywhere within about two to four
days on average; a `low` day takes four to five and a half. Low effort is the
state this panel visits reluctantly, not a basin it falls into. The
stationary distribution — 24% low, 34% moderate, 42% high — is where the
chain settles regardless of its start.

## Persistence, return, and sojourn

`markov_stability()` summarizes each state's dynamics in one row: the
self-transition probability, the stationary share, the mean recurrence time,
and the mean sojourn (how long a visit lasts once entered).

```{r markov-stability, eval=has_nestimate}
effort_spells <- markov_stability(effort_model)

effort_spells
```

```{r markov-stability-plot, eval=has_nestimate, fig.width=7, fig.height=4, fig.alt="Bar panels of persistence, stationary probability, return time, and sojourn time for the three states."}
plot(effort_spells)
```

A high-effort spell lasts 2.4 days on average and recurs every 2.4; a
low-effort spell lasts 1.7 days and recurs only every 4.1. Persistence and
prevalence line up here, but they are logically distinct columns — a rare
state can still be sticky.

## Is the chain well-behaved?

`chain_structure()` reports the Markov-chain diagnostics that other verbs
quietly assume: irreducibility (every state reachable from every other),
aperiodicity, and reversibility.

```{r chain, eval=has_nestimate}
summary(chain_structure(effort_model))
```

The chain is irreducible, aperiodic, and regular, so the stationary
distribution above is well-defined and unique. It is *not* reversible: the
forward process is statistically distinguishable from its time reversal,
which is itself a substantive fact about self-regulation — the routes up the
effort scale and the routes down are not mirror images.

# How predictable is the process?

`transition_entropy()` gives the point estimate; `entropy_bayes()` adds a
posterior credible interval by placing a Dirichlet prior on each row.

```{r entropy-bayes, eval=has_nestimate}
set.seed(2026)
effort_entropy <- entropy_bayes(effort_model)

effort_entropy
```

```{r entropy-bayes-plot, eval=has_nestimate, fig.width=7, fig.height=4, fig.alt="Posterior distributions of the entropy rate, stationary entropy, and redundancy, each a narrow density around its point estimate."}
plot(effort_entropy)
```

The entropy rate is 1.478 bits against a ceiling of `log2(3) = 1.585` — 0.93
on the normalised scale — with a credible interval three hundredths wide:
with 5,600 daily reports, the *uncertainty* about how unpredictable the
process is, is itself small. The redundancy of 0.073 bits is the gap between
the marginal entropy and the conditional one: knowing today's effort removes
under 5% of tomorrow's uncertainty. Daily effort is mostly noise around a
weak first-order signal, and every edge below should be read with that in
mind.

`entropy_network()` maps the same quantity onto the edges — each entry is an
edge's contribution to the row's unpredictability — and returns a network of
the same class, so it can be plotted or pruned like any other:

```{r entropy-network, eval=has_nestimate}
entropy_network(effort_model)
```

# Does history beyond one step matter?

The model assumes the next state depends only on the current one.
`markov_order_test()` (Anderson & Goodman, 1957) tests that assumption.

```{r markov-order, eval=has_nestimate}
set.seed(2026)
effort_order <- markov_order_test(effort_model)

effort_order
```

```{r markov-order-plot, eval=has_nestimate, fig.width=7, fig.height=4, fig.alt="Model-selection chart of log-likelihood, AIC, and BIC across Markov orders zero to three, favouring orders above one."}
plot(effort_order)
```

AIC and the permutation LRT select order 3 and BIC selects order 2: yesterday
carries information about tomorrow beyond today. `path_dependence()` says how
much:

```{r path-dependence, eval=has_nestimate}
effort_history <- path_dependence(effort_model)

effort_history
```

```{r path-dependence-plot, eval=has_nestimate, fig.width=7, fig.height=4, fig.alt="Bar chart of the KL divergence contributed by each two-step context, all values below 0.12 bits."}
plot(effort_history)
```

The weighted divergence is 0.038 bits: history shifts the next-step
distribution by a few hundredths of a bit and flips the modal prediction in
only two of nine contexts. This is the pair of numbers that reconciles a
decisively significant order test with the continued use of a first-order
model — detectable is not the same as large. The first-order network remains
a fair one-step summary; it is not the full process.

`path_counts()` is the raw material — every observed two-step path with its
count:

```{r path-counts, eval=has_nestimate}
path_counts(effort_model)
```

`high → high` alone is 24.0% of all transitions. And when the order test
demands more than one step, `build_hon()` estimates a higher-order network
under its own per-node promotion rule:

```{r hon, eval=has_nestimate}
hon <- build_hon(effort_model, max_order = 2)

pathways(hon)
```

HON promotes no node here — no single context shifts the next-step
distribution enough to justify splitting a state — which is why `pathways()`
comes back empty. Two criteria disagreeing is a result, not a bug: the
likelihood test sees aggregate improvement, HON asks a stricter per-node
question. Report both.

# Simplifying the model

`net_prune()` removes weak edges and returns the same class;
`net_pruning_details()` is the audit trail; `net_deprune()` undoes it.

```{r prune, eval=has_nestimate}
pruned <- net_prune(effort_model, method = "threshold", threshold = 0.3)

net_pruning_details(pruned)
```

Four edges fall, and they are a coherent four: every descending route
(`moderate → low`, `high → low`, `high → moderate`) plus the direct
`low → high` leap. What survives is the ascending ladder and the persistence
loops. Because pruning is recorded rather than destructive, the original
model is one call away:

```{r deprune, eval=has_nestimate}
net_deprune(pruned)
```

# How sure are we about the estimates?

Nestimate offers several answers resting on different machinery: sequence
resampling, node resampling, case dropping, sample splitting, and a
posterior.

## Sequence bootstrap

`bootstrap_network()` resamples the 36 sequences and re-estimates.

```{r bootstrap, eval=has_nestimate}
effort_boot <- bootstrap_network(effort_model, iter = 200, seed = 2026)

effort_boot
```

`summary()` is the tidy one-row-per-edge table:

```{r bootstrap-summary, eval=has_nestimate}
head(summary(effort_boot), 3)
```

Five of nine edges survive — the three persistence loops and the two
ascending steps — while the descending routes move too much across resamples
of 36 students to clear the stability criterion.

## Bayesian edge certainty

`certainty()` reaches its verdict from a Dirichlet posterior on each row,
with no resampling.

```{r certainty, eval=has_nestimate}
set.seed(2026)
certainty(effort_model)
```

All nine edges are posterior-certain, with intervals a few hundredths wide.
The two verbs disagree because they ask different questions: `certainty()`
asks whether an edge's posterior excludes instability given the pooled
transitions, while the bootstrap asks whether the estimate survives swapping
*students* in and out. An edge can be precisely estimated from the pooled
data and still depend on who is in the panel — the four descending edges are
exactly that kind.

## Whole-network statistics: the vertex bootstrap

`vertex_bootstrap()` implements the Snijders–Borgatti (1999) vertex
bootstrap for network-level statistics.

```{r vertex-boot, eval=has_nestimate}
vertex_bootstrap(effort_model, iter = 100, seed = 2026)
```

The mean edge weight carries an interval of [0.23, 0.30] because the vertex
bootstrap resamples *nodes* and there are only three — a deliberately
conservative design for small node sets.

## Case dropping, in two flavours

`centrality_stability()` tracks the centrality *ordering* as sequences are
dropped; `casedrop_reliability()` tracks the edge weights themselves. Both
report the CS-coefficient of Epskamp, Borsboom & Fried (2018).

```{r stability, eval=has_nestimate}
set.seed(2026)
centrality_stability(effort_model, iter = 100)
```

```{r casedrop, eval=has_nestimate}
effort_reliability <- casedrop_reliability(effort_model, iter = 100,
                                           seed = 2026)

effort_reliability
```

```{r casedrop-plot, eval=has_nestimate, fig.width=7, fig.height=4, fig.alt="Case-dropping stability curve showing the correlation between full-sample and subsampled edge weights declining as larger fractions of the 36 sequences are dropped."}
plot(effort_reliability)
```

The verdicts are moderate across the board: the in-strength ordering holds to
a 0.60 drop, the out-strength ordering only to 0.40, and the edge weights sit
exactly at the 0.50 bar. This is what 36 sequences buy, and it is the
sample-size counterweight to the posterior certainty above: report orderings
and intervals, not bare point values.

## Split-half reliability

```{r reliability, eval=has_nestimate}
set.seed(2026)
network_reliability(effort_model)
```

Randomly halving the panel 1,000 times yields two models that correlate at
0.85 on average with a mean absolute edge difference of 0.06 — real
agreement, with real room for movement. All four verbs in this section tell
the same story at different grain: the backbone of the model is solid, the
weaker edges are not yet pinned down.

# Comparing two models

Any two models on the same alphabet can be compared. The pair here splits
the panel at occasion 78 — the first and second halves of the course — with
the same fixed `breaks`, so both models sit on one alphabet, and the same 36
students on both sides.

```{r halves, eval=has_nestimate}
first_half <- ts_tna(
  subset(students, day <= 78),
  value = "effort", id = "name", time = "day",
  discretization = "threshold", breaks = c(40, 70),
  labels = c("low", "moderate", "high")
)

second_half <- ts_tna(
  subset(students, day > 78),
  value = "effort", id = "name", time = "day",
  discretization = "threshold", breaks = c(40, 70),
  labels = c("low", "moderate", "high")
)
```

## The difference itself

`subtract_networks()` returns the edge-by-edge difference as an object
cograph can draw:

```{r subtract, eval=has_nestimate}
half_difference <- subtract_networks(first_half, second_half)

half_difference
```

```{r subtract-plot, eval=has_nestimate && has_cograph, fig.width=6.5, fig.height=5, fig.alt="Difference network of the two course halves, edges coloured by the sign of the change and all magnitudes small."}
cograph::plot_difference(half_difference)
```

The largest gap on any edge is 0.033 (`low → moderate`).

## Four more verdicts on the same pair

`compare_model()` condenses the pair into a metric panel; `bayes_compare()`
asks whether each edge difference is credibly nonzero; `vertex_compare()`
tests the network-level statistics; and the workflow vignette's
`permutation()` run supplies the sequence-level test.

```{r compare, eval=has_nestimate}
half_panel <- compare_model(first_half, second_half)

half_panel
```

```{r compare-plot, eval=has_nestimate, fig.width=7, fig.height=4.5, fig.alt="Comparison panel chart of the two half-course models across weight-deviation, correlation, and similarity metrics, all indicating near-identity."}
plot(half_panel)
```

```{r bayes-compare, eval=has_nestimate}
set.seed(2026)
bayes_compare(first_half, second_half)
```

```{r vertex-compare, eval=has_nestimate}
vertex_compare(first_half, second_half, iter = 200, seed = 2026)
```

Five verbs, one verdict. The matrices correlate at 0.990 with rank and sign
agreement of 1; the posterior finds zero of nine edges credibly different;
the vertex comparison finds no network-level statistic moved (all
p ≥ 0.67); and the permutation test in the workflow vignette finds nothing
at the sequence level either. The course's second half reproduces its first,
in level and in dynamics. When the verbs *disagree* — a posterior flagging
tiny-but-credible differences that a sequence-level permutation absorbs —
report the magnitude next to whichever test is used; the grouped
weekday/weekend comparison in `vignette("group-models", package = "tsn")`
shows what a real, localized difference looks like on these same students.

# Link prediction

`predict_links()` scores unobserved dyads by six structural predictors:

```{r predict, eval=has_nestimate}
predict_links(effort_model)
```

It runs, but a three-state transition network is fully connected between
states, which leaves link prediction nothing to rank; it earns its place on
larger, pruned alphabets, where a missing edge is a real hypothesis.

# Grouped models

When the comparison variable is a column of the data,
`ts_tna(group = ...)` builds one network per level and returns a Nestimate
`netobject_group`. The grouped verbs verified against it are
`net_centrality()`, `state_distribution()`, `transition_entropy()`,
`passage_time()`, `certainty()`, `net_prune()`, `compare_model()`,
`permutation()`, `bootstrap_network()`, `markov_order_test()`,
`centrality_stability()`, `casedrop_reliability()`, and `mosaic_plot()`;
`summary()` and `as.data.frame()` return tidy per-group tables, and
`plot()` draws one selected group. `net_edge_betweenness()` also runs, but
its result drops back to a plain `netobject_group`. The dedicated vignette
`vignette("group-models", package = "tsn")` works through all of it on the
`esm_srl` weekday/weekend split.

# What does not apply, and why

Three families of Nestimate verbs reject a transition model, each for a
principled reason rather than a missing feature:

* **Sequence-data verbs.** `frequencies()`, `mark_first_state()`,
  `mark_terminal_state()`, `entropy_trajectory()`, `mosaic_analysis()`,
  `plot_mosaic()`, and `as_htna()` operate on a raw wide data frame of
  sequences, not on an estimated model. They belong *before* model
  construction in a workflow, applied to sequence data you hold yourself.
* **Psychometric-network verbs.** `predictability()` and the
  `glasso`/`pcor`/`cor` estimator family need a precision or correlation
  structure estimated from multivariate data. A transition matrix is a
  conditional-probability object, not a conditional-independence one; the
  quantities these verbs compute are undefined for it.
* **Verbs that need extra structure.** `cluster_network()` and
  `build_clusters()` require a chosen number of clusters and are aimed at
  multi-sequence typologies; `wtna()` needs one-hot action data;
  `sequence_compare()` wants sequence data plus a grouping vector. And
  `as_tna()` is unnecessary by design — a tsn model needs no conversion,
  which is the point of this vignette.

# The compatibility map

| Question | Verb | Tidy access | Plot | Notes |
|---|---|---|---|---|
| What are the estimates? | `as.data.frame()`, `as.matrix()`, `extract_transition_matrix()`, `extract_initial_probs()` | direct | via `plot()` on the model | tsn accessors include the diagonal |
| Between-state edges only | `extract_edges()` | direct | — | **excludes self-transitions** |
| Marginal composition | `state_distribution()`, `state_frequencies()` | data frame | `mosaic_plot()`, `plot_state_frequencies()` | |
| Composition over time / carpet | `distribution_plot()`, `sequence_plot()` | — | draws directly | stacked area; sequence heatmap |
| State/edge importance | `net_centrality()`, `net_edge_betweenness()` | data frame / network | `plot()` on both | loops excluded from centrality by default |
| Travel times between states | `passage_time()` | print | `plot()` | Kemeny & Snell (1976) |
| Persistence, recurrence, sojourn | `markov_stability()` | `$stability` table | `plot()` | |
| Chain diagnostics | `chain_structure()` | `summary()` | `plot()` heatmap | irreducibility, reversibility |
| Predictability | `transition_entropy()`, `entropy_bayes()`, `entropy_network()` | print / `summary()` | `plot()` on both entropy results | rate in bits; ceiling `log2(n_states)` |
| History beyond one step | `markov_order_test()`, `path_dependence()`, `path_counts()`, `build_hon()` + `pathways()` | print / data frame | `plot()` on test and dependence | test the size, not only the p |
| Simplification | `net_prune()`, `net_pruning_details()`, `net_deprune()` | same class back | via model `plot()` | prune after testing |
| Edge uncertainty | `bootstrap_network()`, `certainty()` | `summary()` | — | resampling vs Bayesian; they can disagree |
| Network-level uncertainty | `vertex_bootstrap()` | data frame | `plot()` | Snijders & Borgatti (1999) |
| Sample-size reliability | `centrality_stability()`, `casedrop_reliability()`, `network_reliability()` | print | `plot()` on all three | CS-coefficient bar: 0.5 / 0.7 |
| Two-model comparison | `permutation()`, `compare_model()`, `subtract_networks()`, `bayes_compare()`, `vertex_compare()` | `summary()` / data frame | `plot()`; `cograph::plot_difference()` | report magnitudes with tests |
| Link prediction | `predict_links()` | print | — | needs a sparse alphabet to be informative |
| Many-group comparison | `ts_tna(group=)` + grouped verbs | `as.data.frame()` | `plot(x, group=)`, `mosaic_plot()` | see `vignette("group-models")` |
| Rendering | `plot()`, `cograph::splot()` | — | — | requires cograph |

# References

Anderson, T. W., & Goodman, L. A. (1957). Statistical inference about
Markov chains. *The Annals of Mathematical Statistics*, 28(1), 89–110.

Epskamp, S., Borsboom, D., & Fried, E. I. (2018). Estimating psychological
networks and their accuracy: A tutorial paper. *Behavior Research Methods*,
50(1), 195–212.

Kemeny, J. G., & Snell, J. L. (1976). *Finite Markov Chains*. Springer.

Snijders, T. A. B., & Borgatti, S. P. (1999). Non-parametric standard
errors and tests for network statistics. *Connections*, 22(2), 161–170.

Xu, J., Wickramarathne, T. L., & Chawla, N. V. (2016). Representing
higher-order dependencies in networks. *Science Advances*, 2(5), e1600028.
