---
title: "Transition Networks from Time Series: Estimation and Inference with Nestimate"
author:
  - name: "Mohammed Saqr"
    affiliation: "University of Eastern Finland"
  - name: "Sonsoles López-Pernas"
    affiliation: "University of Eastern Finland"
  - name: "Manuel J. Gómez"
    affiliation: "University of Murcia"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Transition Networks from Time Series: Estimation and Inference with Nestimate}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 150,
  out.width = "100%"
)
library(tsn)
has_nestimate <- requireNamespace("Nestimate", quietly = TRUE)
has_cograph <- requireNamespace("cograph", quietly = TRUE) &&
  requireNamespace("igraph", quietly = TRUE)
```

# Introduction

A numeric time series records the level of a quantity over time. A transition
network records something different: how a system moves between a small number
of qualitative *states*, with nodes for the states and directed edges for the
probability of passing from one to the next. The two descriptions answer
different questions. The series says how high a measurement was; the network
says what a measurement is likely to be followed by. Recovering the second from
the first takes two steps, reducing the continuous series to a sequence of
states and estimating the state-to-state transitions, and then a third that is
easy to skip: deciding what such a model may legitimately claim.

The **tsn** package performs the first two steps in a single call and hands the
result to **Nestimate**, which supplies the third. `ts_tna()` discretizes a
series, estimates the transition network, and returns a native Nestimate model,
so Nestimate's descriptive and inferential verbs apply to it with no conversion
step in between. This vignette follows one analysis through the whole path:
building the model, reading its structure, and, where the data support it,
testing that structure with bootstrap confidence intervals, centrality
stability, a Markov-order test, and a permutation comparison of two periods.

The distinction between description and inference organizes everything below.
Descriptive summaries such as centralities, state distributions, entropy, and
pruning ask what a fitted network looks like, and a network fitted to a single
series answers them. Inference asks how far the estimate would move under a
different sample, and a single series is a single sample: there is nothing to
resample. The dividing line is the unit of analysis, and knowing where it falls
is what keeps a transition model from being read for more than it can support.

```{r libs, eval=has_nestimate}
library(tsn)
library(Nestimate)

data(motivation)
data(steps)
```

# From measurements to a model

`ts_tna()` takes a data frame and the name of one numeric column, cuts that
column into states, and estimates the transition network in one call. Here it
runs on the first 200 observations of the `pleasure` rating from the
`motivation` study, an intensive-longitudinal record of momentary experience.

```{r build, eval=has_nestimate}
pleasure <- head(motivation, 200)

model <- ts_tna(
  pleasure,
  series = "pleasure",
  labels = c("low", "mid", "high")
)

model
```

The printed object is the complete model: a row-stochastic weight matrix whose
entry *(i, j)* is the probability of moving to state *j* given that the system
is currently in state *i*, together with the initial-state distribution. Each
row sums to one because from any state the process must go somewhere.

## The bridge: this object *is* a Nestimate model

Everything that follows rests on one property of the returned object.

```{r class, eval=has_nestimate}
class(model)
```

`ts_tna()` does not return a private structure that must be converted before
Nestimate can read it. It returns an object that inherits from **`netobject`**,
Nestimate's core network class, and Nestimate's verbs dispatch on that class
directly. There is no export step and no adapter call between the model and the
tests applied to it, which is what lets a single analysis move from
construction to inference without leaving the object behind.

## Four constructors, one interface

The same discretized sequence supports four definitions of an edge, and the four
constructors take identical arguments and return the same class. `ts_tna()`
gives relative frequencies, so an edge is the conditional probability of the
next state. `ts_ftna()` gives the raw transition counts behind those
probabilities. `ts_atna()` applies attention weighting, and `ts_cna()` counts
undirected co-occurrence rather than directed movement. Changing the edge
definition means changing only the constructor.

```{r builders-run, eval=has_nestimate}
ts_ftna(pleasure, series = "pleasure", labels = c("low", "mid", "high"))
```

The remainder of the vignette uses `ts_tna()`, because the inferential tests are
written for probabilities.

# One series, and what it can support

Asking a single-series model for a bootstrap returns a warning rather than a
result:

> A network with one long sequence is not recommended and can't be validated
> using bootstrap and other confirmatory testings.

The warning is correct. A bootstrap resamples *sequences* to estimate how far a
quantity would move under a different sample of them; one series is one
sequence, so there is nothing to resample and no sampling distribution to form.
The permutation and stability tests rest on the same logic.

This is a statement about the unit of analysis, not a dead end. A long series
contains many transitions; what it lacks is a way to group them into units that
can be resampled. The `segment` argument supplies that unit by cutting the
series into shorter blocks, each treated as its own sequence. This is the block
bootstrap, and tsn provides its two standard forms.

The series below is one participant's 299 daily step counts. Two features of the
call matter before the segmenting itself. States are cut at 5,000 and 10,000
steps with `discretization = "threshold"` rather than at the default tertiles,
so the states mean the same thing outside this sample; the next section returns
to why that matters. And segmentation is applied *after* discretization, so
every block inherits the same cut points and no block is ever discretized
against itself.

```{r segment-setup, eval=has_nestimate}
walkers <- subset(steps, !is.na(steps))
one_walker <- subset(walkers, id == 193)

whole <- ts_tna(
  one_walker,
  value = "steps",
  time = "day",
  discretization = "threshold",
  breaks = c(5000, 10000),
  labels = c("sedentary", "moderate", "active")
)
```

Partitioning cuts the 299 observations into consecutive, non-overlapping blocks.
At `segment = 10` there are thirty of them.

```{r segment-blocks, eval=has_nestimate}
blocks <- ts_tna(
  one_walker,
  value = "steps",
  time = "day",
  segment = 10,
  discretization = "threshold",
  breaks = c(5000, 10000),
  labels = c("sedentary", "moderate", "active")
)

bootstrap_network(blocks, iter = 500, seed = 2024)
```

Thirty sequences, and the bootstrap now runs. Partitioning is not free: every
cut destroys the transition that straddles it, so thirty blocks cost roughly
thirty of the 298 transitions and the estimate shifts slightly.

Sliding the window instead of partitioning keeps all of them. At `segment = 2,
overlap = TRUE` the blocks are the consecutive lag-1 pairs, so every transition
appears exactly once and the network is *identical* to the unsegmented one.

```{r segment-lag1, eval=has_nestimate}
lag_one <- ts_tna(
  one_walker,
  value = "steps",
  time = "day",
  segment = 2,
  overlap = TRUE,
  discretization = "threshold",
  breaks = c(5000, 10000),
  labels = c("sedentary", "moderate", "active")
)

all.equal(as.matrix(lag_one), as.matrix(whole))
```

```{r segment-lag1-boot, eval=has_nestimate}
bootstrap_network(lag_one, iter = 500, seed = 2024)
```

The two schemes trade precision against independence. Sliding blocks preserve
the point estimate exactly, yield the most units, and tend to give narrower
intervals: a mean width of 0.18 here against 0.21 for the partitioned blocks,
narrow enough to call a third edge significant where partitioning calls two.
The tendency is not a guarantee, since two of the nine edges come out wider. And
the narrowness carries a cost. Overlapping blocks share observations, so they
are not independent, and resampling lag-1 pairs assumes exactly the first-order
Markov property the model itself assumes. That is the assumption the
Markov-order test below rejects for these data, which makes those intervals
optimistic.

Partitioned blocks are the more conservative choice. A block preserves whatever
dependence exists inside it, so structure beyond one lag survives resampling,
at the price of the boundary transitions lost to the cuts. Blocks of 10 to 30
retain 90% to 97% of the transitions, which is usually the right trade; blocks
of 2 retain only half and are better avoided.

None of this manufactures information. Segmenting a single series asks how stable
one person's own dynamics are, and says nothing about anyone else. A claim about
people in general needs people in general, so the rest of the vignette uses all
151 of them.

## Choosing states that mean something

The default `discretization = "quantile"` cuts at the tertiles, forcing each of
the three states to hold a third of the observations, so the state sizes carry
no information, by construction. That is why every model built from the `steps`
data uses fixed thresholds of 5,000 and 10,000 steps instead. Fixed breaks make
the states interpretable outside the sample, and they give every model one
shared alphabet, which is what makes the period comparison below legitimate.
(The grouped `pleasure` model at the end keeps the quantile default; grouping
pools the series before cutting it, so its groups share an alphabet without
breaks being named.)

```{r many, eval=has_nestimate}
walk_model <- ts_tna(
  walkers,
  value = "steps",
  id = "id",
  time = "day",
  discretization = "threshold",
  breaks = c(5000, 10000),
  labels = c("sedentary", "moderate", "active")
)

walk_model
```

The states now mean something outside the data, the same cut points apply to
every comparison made later, and the state distribution is worth reading.

```{r distribution, eval=has_nestimate}
state_distribution(walk_model)
```

Every state is strongly self-preferring, and an active day is the most
persistent of all at 0.62: activity, once reached, tends to continue. Step
counts show marked day-to-day inertia.

## One network per person

The pooled model estimates a single network across everyone. `series_networks()`
splits it into one complete model per participant, each rebuilt on the same
alphabet so node sets stay aligned and comparable.

```{r per-person, eval=has_nestimate}
per_person <- series_networks(walk_model)

head(summary(per_person))
```

Each row indexes a full `ts_tna` model that prints and plots on its own: the
models are the list elements, and this is the tidy index over them. Because the
cut points are fixed step counts rather than sample quantiles, a participant's
personal network describes their own activity, not their rank within the sample.

# Describing the model

## Centrality

```{r centrality, eval=has_nestimate}
net_centrality(walk_model)
```

The message notes that centralities exclude self-transitions by default, which
matters for a transition network: the diagonal is usually the largest entry, so
a state can be highly persistent while receiving little traffic from elsewhere.
Excluding the diagonal measures how central a state is to *movement* between
states rather than to staying put. Passing `loops = TRUE` includes it.

## How much of the sequence is structure, and how much is noise

```{r entropy, eval=has_nestimate}
transition_entropy(walk_model)
```

The entropy rate is normalized against `log2(n_states)`, so 0 is a perfectly
deterministic process and 1 means every next state is equally likely from
wherever the process currently sits. A value of 1 requires more than an
uninformative current state: a lopsided but state-independent process still
scores well below it. Cut at the tertiles, these same step counts score 0.93;
cut at 5,000 and 10,000 they score 0.85. The cut points are not cosmetic: they
decide how much structure is left to find.

# Pruning

Not every edge deserves a place in the picture. `net_prune()` returns a model of
the same class with weak edges removed, so its result flows straight back into
any verb above.

```{r prune, eval=has_nestimate}
net_prune(walk_model, method = "threshold", threshold = 0.3)
```

Three edges fall below the threshold: every route *into* sedentary from a higher
state, together with the direct leap from sedentary to active. What survives
says that a sedentary day is harder to fall into than to climb out of. The
`sedentary -> moderate` edge is a strong 0.445, while nothing descends into
sedentary above 0.3.

Prune after testing rather than before, however. A threshold ranks edges by
size, and size is not the same as interest: a small but reliably estimated edge
can matter more than a large but uncertain one.

# Bootstrap: which edges survive resampling?

`bootstrap_network()` resamples the sequences with replacement, re-estimates the
network on each resample, and reports a confidence interval for every edge.

```{r bootstrap, eval=has_nestimate}
boot <- bootstrap_network(walk_model, iter = 200, seed = 2024)

boot
```

`summary()` returns the tidy, one-row-per-edge table.

```{r bootstrap-summary, eval=has_nestimate}
summary(boot)
```

With 151 sequences behind it, every edge is estimated precisely and all nine
survive resampling. Confidence intervals a few hundredths wide are what a large
sample buys, and they are the ground on which the descriptive reading above can
be reported rather than merely observed.

# Stability: would the same states rank first in a smaller study?

`centrality_stability()` repeatedly drops a proportion of the sequences,
re-computes the centralities, and reports the largest proportion that can be
dropped while the ordering still tracks the original, a correlation-stability
coefficient.

```{r stability, eval=has_nestimate}
centrality_stability(walk_model, iter = 100)
```

All three coefficients reach 0.90, against conventional bars of 0.5 for
acceptable and 0.7 for excellent. That measures how stable the *ordering* is
under resampling; it does not speak to whether the ordering is substantively
meaningful, only to whether a smaller study would have recovered it.

# Markov order: is a first-order model enough?

Every transition network here assumes the Markov property, that the next state
depends on the current one and nothing earlier. That assumption is testable, by
comparing the fit of models of increasing order.

```{r markov, eval=has_nestimate}
markov_order_test(walk_model)
```

The test disagrees with the first-order model. AIC, BIC, and the permutation
likelihood-ratio test all prefer a higher order, meaning yesterday carries
information about tomorrow beyond what today already conveys. This does not
invalidate the first-order network, but it bounds the claim: the network is a
summary of one-step behaviour, not a full description of the process.

The natural follow-up is a higher-order network.

```{r hon, eval=has_nestimate}
build_hon(walk_model, max_order = 2)
```

The higher-order construction keeps every node at order 1: it finds no
individual context that shifts the next-step distribution enough to justify
splitting the node. The two criteria disagree because they ask different
questions. The likelihood test detects aggregate improvement pooled across all
sequences, while the higher-order rule applies a stricter, per-node test. Both
results are real, and both are worth reporting: there is higher-order structure
in aggregate, but not localized in any single state.

# Comparing two periods

The tests so far describe one network. `permutation()` compares two, shuffling
the group labels to build a null distribution for every edge. The `steps` series
runs from April 2019 to March 2020; splitting at the start of November contrasts
the warmer months with the colder ones. The two spans are unequal in length,
roughly six and a half months against four.

Both models are built with the same fixed `breaks`, which matters more than it
appears: had each period been discretized on its own quantiles, the two networks
would rest on different cut points, and any difference between them could reflect
the thresholds rather than behaviour.

```{r two-models, eval=has_nestimate}
cutoff <- as.Date("2019-11-01")

early <- ts_tna(
  subset(walkers, as.Date(day) < cutoff),
  value = "steps",
  id = "id",
  time = "day",
  discretization = "threshold",
  breaks = c(5000, 10000),
  labels = c("sedentary", "moderate", "active")
)

late <- ts_tna(
  subset(walkers, as.Date(day) >= cutoff),
  value = "steps",
  id = "id",
  time = "day",
  discretization = "threshold",
  breaks = c(5000, 10000),
  labels = c("sedentary", "moderate", "active")
)

comparison <- permutation(early, late, iter = 2000, seed = 2024)

comparison
```

```{r perm-table, eval=has_nestimate}
summary(comparison)
```

None of the nine edge comparisons reaches significance; the smallest p-value is
0.07. The estimates are not identical, but no difference survives the
permutation null. That is a more informative result than a scattering of small
p-values would have been, because the two periods differ in how much people
walked in the first place:

```{r composition, eval=has_nestimate}
state_distribution(early)

state_distribution(late)
```

Active days fall from 46.9% before November to 42.8% after it, and sedentary
days rise. The seasonal shift is therefore plain in the *composition* of the
states while no comparison of the *transitions* between them reaches
significance. The estimated sedentary-to-moderate probability, for instance, is
0.446 before November and 0.444 after, with a permutation p-value of 0.94.

Two cautions bound how far that carries. Failing to reject nine tests is not the
same as showing the dynamics are identical, and the two models do not rest on
quite the same people: 140 participants contribute before November and 149
after, with 138 in both.

That distinction is the reason to build a transition network at all. A
comparison of means or of state proportions finds the seasonal difference and
stops there; the transition model asks a further question of the same
population, whether the *dynamics* moved with the season, and finds no evidence
that they did.

# More than two groups: one model per context

Building two models by hand is manageable when there are two. When the variable
that separates the groups is already a column in the data, the `group` argument
builds one network per level in a single call, and it scales past two, which the
pairwise form does not.

The `motivation` study records where each measurement was taken in
`task_context_type`, so the pleasure ratings can be split by context.

```{r grouped, eval=has_nestimate}
by_context <- ts_tna(
  motivation,
  series = "pleasure",
  group = "task_context_type",
  labels = c("low", "mid", "high")
)

by_context
```

Two things happen here that would be tedious to arrange by hand. The states are
cut from the *pooled* series before the split, so all four networks rest on one
set of thresholds and share a node set, the same discipline the fixed `breaks`
enforced earlier, applied automatically. And the result is a Nestimate
`netobject_group`, so the compatible grouped verbs take it directly:
`net_prune()`, `state_distribution()`, `net_centrality()`, `compare_model()`,
`permutation()`, `mosaic_plot()`, and their neighbours. Not every verb applies:
those needing a precision matrix or a clustering attribute reject a transition
model, and there is no grouped `plot()` method, so networks are plotted one at a
time.

`as.data.frame()` gives the tidy view of how much data stands behind each
network:

```{r grouped-index, eval=has_nestimate}
as.data.frame(by_context, what = "groups")
```

The sequence counts repay a second look: `Home` has 1,324 observations but 832
sequences. That is not padding. A run is cut at every change of series *and* at
every change of group, and the two cuts are not the same. A change of series
separates measurements that were never consecutive; a change of context falls
between two genuinely adjacent measurements, and the transition across it is
excluded from both networks because it belongs to neither context. This study
alternates context from one measurement to the next (4,871 observations form
2,632 contiguous runs), so each unbroken stretch within a context becomes its
own sequence, and a pair of measurements separated by an intervening visit to a
different context is never counted as a transition, because it never was one.

That rule costs nothing when a group is constant within a series, which is the
usual case for a condition or a cohort: each series then contributes one
sequence and keeps its own ID.

```{r grouped-distribution, eval=has_nestimate}
state_distribution(by_context)
```

The contexts are not alike. Pleasure is low on 62.5% of measurements taken at
home and high on only 6.6% of them; personal time reverses that ordering: high
on 45.6%, low on 19.3%. Work sits between the two on all three states. A mosaic
plot shows the same contrast in the transition structure at a glance, one panel
per context:

```{r grouped-mosaic, eval=has_nestimate, fig.width=9, fig.height=6, fig.alt="Mosaic panels of the four context-specific transition networks, in which tile area encodes transition probability from each current state to each next state, showing the home context concentrated in the low-to-low cell and personal time in the high-to-high cell."}
mosaic_plot(by_context)
```

Tile area encodes the transition probability from a current state (rows) to a
next state (columns). The home panel is dominated by its low-to-low cell and the
personal panel by its high-to-high cell, so the difference in *level* between the
contexts reappears as a difference in *persistence*: wherever pleasure tends to
be, it tends to stay.

Because every network shares an alphabet, two of them can be set side by side
and selected by name:

```{r grouped-compare, eval=has_nestimate}
compare_model(by_context, i = "Home", j = "Work")
```

Pruning applies group-wise and returns a grouped model, so the collection can be
carried through a workflow without being taken apart:

```{r grouped-prune, eval=has_nestimate}
net_prune(by_context, method = "threshold", threshold = 0.3)
```

The index table states one caution plainly rather than hiding it: `Other` holds
three observations and two sequences. It is a level of the column, so it gets a
network, but a network estimated from a single transition is not a result. Read
the `sequences` and `observations` columns before reading the edges.

# Plotting a single model

A grouped model has no `plot()` method, but any single network does, including
the pooled step model. The plot draws the three states as nodes and the
transition probabilities as directed edges, with self-loops for persistence.

```{r plot, eval=has_nestimate && has_cograph, fig.width=9, fig.height=6, fig.alt="Transition network of daily step-count states, with sedentary, moderate, and active nodes joined by directed edges whose thickness reflects transition probability and self-loops that mark each state's persistence."}
plot(walk_model)
```

The heavy self-loops confirm what the state distribution reported numerically:
each state mostly returns to itself, and movement between states is the smaller
part of the process. The plotting vignette covers the plot types and their
arguments in full.

# Where the line falls

Everything descriptive in this vignette, from centralities, state distributions,
entropy, and pruning to per-participant splits and the grouped models, runs on a
model built from a single series. Everything inferential, meaning the bootstrap,
stability, the Markov-order test, and the permutation test, requires many
sequences, because resampling and permutation both treat the sequence as the
unit of analysis.

Three practices follow. Build with `id` from the start if you intend to test
anything, so the sequences exist when a test asks for them. Fix your `breaks`
before you compare anything, or pass `group` and let the pooled discretization
fix them for you, so a difference between networks is a difference in behaviour
and not in thresholds. And when a comparison rests on few sequences, say so: the
`sequences` column exists to be read.
