---
title: "Building a Model from a Time Series and Testing It with Nestimate"
author: "tsn package authors"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Building a Model from a Time Series and Testing It with Nestimate}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 150,
  out.width = "100%"
)
library(tsn)
has_nestimate <- requireNamespace("Nestimate", quietly = TRUE)
has_cograph <- requireNamespace("cograph", quietly = TRUE) &&
  requireNamespace("igraph", quietly = TRUE)
```

# What this vignette does

A time series is a sequence of numbers. A transition network is a model of how a
system moves between *states*. This vignette covers the whole distance between
them: it turns a series into states, estimates a transition-network model, and
then submits that model to the inferential machinery in **Nestimate** —
bootstrap confidence intervals, centrality stability, a Markov-order test, and a
permutation test comparing two periods.

```{r libs, eval=has_nestimate}
library(tsn)
library(Nestimate)

data(motivation)
data(steps)
```

# From measurements to a model

`ts_tna()` takes a data frame and the name of one numeric column, cuts that
column into states, and estimates the transition network in a single call. Here
it runs on the first 200 observations of the `pleasure` rating from the
`motivation` study.

```{r build, eval=has_nestimate}
pleasure <- head(motivation, 200)

model <- ts_tna(
  pleasure,
  series = "pleasure",
  labels = c("low", "mid", "high")
)

model
```

The printed object is the whole model: a row-stochastic weight matrix in which
entry *(i, j)* is the probability of moving to state *j* given that the system is
currently in state *i*, together with the initial-state distribution.

## The bridge: this object *is* a Nestimate model

Everything that follows rests on one fact.

```{r class, eval=has_nestimate}
class(model)
```

`ts_tna()` does not return a private structure that has to be converted. It
returns an object inheriting from **`netobject`**, which is Nestimate's core
network class, and Nestimate's verbs dispatch on that class. There is no export
step and no adapter call in between.

## Four constructors, one interface

The same discretized sequence supports four definitions of an edge, and all four
constructors take identical arguments and return the same class. `ts_tna()` gives
relative frequencies, so an edge is the probability of the next state.
`ts_ftna()` gives the raw transition counts behind those probabilities.
`ts_atna()` applies attention weighting, and `ts_cna()` counts undirected
co-occurrence rather than directed movement.

Swapping the constructor is the only change required.

```{r builders-run, eval=has_nestimate}
ts_ftna(pleasure, series = "pleasure", labels = c("low", "mid", "high"))
```

The rest of this vignette uses `ts_tna()`, because probabilities are what the
inferential tests are written for.

# One series, and what it can support

Ask a single-series model for a bootstrap and Nestimate objects:

> A network with one long sequence is not recommended and can't be validated
> using bootstrap and other confirmatory testings.

The warning is correct as stated. A bootstrap resamples *sequences* to ask how
far the estimate would move under a different sample of them, and one series is
one sequence: nothing to resample, no sampling distribution. The permutation and
stability tests rest on the same logic.

That is a statement about the *unit of analysis*, though, not a dead end. A long
series contains many transitions; what it lacks is a way to group them into
units that can be resampled. `segment` supplies that by cutting the series into
shorter blocks, each treated as its own sequence. This is the block bootstrap,
and `tsn` offers both of its standard forms.

The series below is one participant's 299 daily step counts. Two things about
the call are worth noting before the segmenting itself. States are cut at 5,000
and 10,000 steps with `discretization = "threshold"` rather than at the default
tertiles, so that they mean something outside this sample — the next section
returns to why that matters. And segmentation is applied *after* discretization,
so every block inherits the same cut points; blocks are never discretized
against themselves.

```{r segment-setup, eval=has_nestimate}
walkers <- subset(steps, !is.na(steps))
one_walker <- subset(walkers, id == 193)

whole <- ts_tna(
  one_walker,
  value = "steps",
  time = "day",
  discretization = "threshold",
  breaks = c(5000, 10000),
  labels = c("sedentary", "moderate", "active")
)
```

`segment = 10` partitions the 299 observations into consecutive blocks of ten.

```{r segment-blocks, eval=has_nestimate}
blocks <- ts_tna(
  one_walker,
  value = "steps",
  time = "day",
  segment = 10,
  discretization = "threshold",
  breaks = c(5000, 10000),
  labels = c("sedentary", "moderate", "active")
)

bootstrap_network(blocks, iter = 500, seed = 2024)
```

Thirty sequences, and the bootstrap runs without complaint. Partitioning is not
free: every cut destroys the transition that straddles it, so thirty blocks cost
roughly thirty of the 298 transitions and the estimate shifts slightly.

Sliding the window instead keeps all of them. At `segment = 2, overlap = TRUE`
the blocks are the consecutive lag-1 pairs — every transition appears exactly
once, so the network is *identical* to the unsegmented one.

```{r segment-lag1, eval=has_nestimate}
lag_one <- ts_tna(
  one_walker,
  value = "steps",
  time = "day",
  segment = 2,
  overlap = TRUE,
  discretization = "threshold",
  breaks = c(5000, 10000),
  labels = c("sedentary", "moderate", "active")
)

all.equal(as.matrix(lag_one), as.matrix(whole))
```

```{r segment-lag1-boot, eval=has_nestimate}
bootstrap_network(lag_one, iter = 500, seed = 2024)
```

So which to use? The sliding blocks preserve the estimate exactly and give the
most units, and their intervals tend to come out narrower — mean width 0.18
against 0.21 here, narrow enough to call a third edge significant where the
partitioned blocks call two. It is a tendency, not a guarantee: two of the nine
edges come out wider. That narrowness is the catch:
overlapping blocks share observations, so they are not independent, and
resampling lag-1 pairs assumes precisely the first-order Markov property the
model itself assumes. The Markov-order test later in this vignette finds that
assumption does not hold here — so those intervals are optimistic.

Partitioned blocks are the more conservative choice. A block preserves whatever
dependence exists inside it, so structure beyond one lag survives resampling,
and you pay for that in lost boundary transitions. Blocks of 10 to 30 keep 90%
to 97% of the transitions, which is usually the right trade; blocks of 2 keep
only half and should be avoided.

None of this manufactures information. Segmenting a single series lets you ask
how stable that person's own dynamics are, and nothing more — it says nothing
about anyone else. For a claim about people in general you need people in
general, so the rest of this vignette uses all 151 of them.

## Choosing states that mean something

The default `discretization = "quantile"` cuts at the tertiles, which forces the
three states to hold a third of the observations each — so the state sizes carry
no information, by construction. That is why every model built from the `steps`
data uses the fixed 5,000 and 10,000 step thresholds instead. Beyond making the
states interpretable, fixed `breaks` give those models one shared alphabet, which
is what makes the period comparison below legitimate. (The grouped pleasure model
at the end keeps the quantile default; grouping pools the series before cutting
it, so its groups share an alphabet without breaks being named.)

```{r many, eval=has_nestimate}
walk_model <- ts_tna(
  walkers,
  value = "steps",
  id = "id",
  time = "day",
  discretization = "threshold",
  breaks = c(5000, 10000),
  labels = c("sedentary", "moderate", "active")
)

walk_model
```

Now the states mean something outside the data, the same cut points apply to
every comparison made later, and the distribution is worth reading.

```{r distribution, eval=has_nestimate}
state_distribution(walk_model)
```

Every state is strongly self-preferring, and an active day is the most persistent
of all at 0.62. Step counts are sticky.

## One network per person

The pooled model estimates a single network across everyone. `series_networks()`
splits it back into one complete model per participant, each rebuilt on the same
alphabet so node sets stay aligned and comparable.

```{r per-person, eval=has_nestimate}
per_person <- series_networks(walk_model)

head(summary(per_person))
```

Each row indexes a full `ts_tna` model that prints and plots on its own — the
models are the list elements, and this is the tidy index of them. Because the
cut points here are fixed step counts rather than sample quantiles, a
participant's personal network describes their own activity, not their rank
within the sample.

# Describing the model

## Centrality

```{r centrality, eval=has_nestimate}
net_centrality(walk_model)
```

Note the message: centralities exclude self-transitions by default. That matters
for a transition network, where the diagonal is usually the largest entry — a
state can be highly persistent while receiving little traffic from elsewhere.
Pass `loops = TRUE` to include the diagonal.

## How much of the sequence is structure, and how much is noise

```{r entropy, eval=has_nestimate}
transition_entropy(walk_model)
```

The entropy rate is normalised against `log2(n_states)`, so 0 is a perfectly
deterministic process and 1 means every next state is equally likely from
wherever you are — not merely that the current state is uninformative, since a
lopsided but state-independent process scores well below 1. These same step counts cut at the tertiles score 0.93; cut at 5,000 and
10,000 they score 0.85. Cut points are not a cosmetic choice — they decide how
much structure there is left to find.

# Pruning

Not every edge deserves to be in the picture. `net_prune()` returns a model of
the same class with weak edges removed, so it flows straight back into any verb
above.

```{r prune, eval=has_nestimate}
net_prune(walk_model, method = "threshold", threshold = 0.3)
```

Three edges fall below the threshold: every route *into* sedentary from a higher
state, plus the direct leap from sedentary to active. What survives says that a
sedentary day is harder to fall into than to climb out of — `sedentary ->
moderate` is a strong 0.445, while nothing descends into sedentary above 0.3.

Prune after testing rather than before, though: a threshold sorts edges by size,
and size is not the same thing as interest.

# Bootstrap: which edges survive resampling?

```{r bootstrap, eval=has_nestimate}
boot <- bootstrap_network(walk_model, iter = 200, seed = 2024)

boot
```

`summary()` returns the tidy, one-row-per-edge table.

```{r bootstrap-summary, eval=has_nestimate}
summary(boot)
```

With 151 sequences behind it every edge is estimated precisely and all nine
survive. Confidence intervals a few hundredths wide are what a large sample buys.

# Stability: would the same states rank first in a smaller study?

`centrality_stability()` repeatedly drops a proportion of the sequences,
re-computes the centralities, and reports the largest proportion that can be
dropped while the ordering still tracks the original.

```{r stability, eval=has_nestimate}
centrality_stability(walk_model, iter = 100)
```

All three coefficients come out at 0.90, against a conventional bar of 0.5 for
acceptable and 0.7 for excellent. That speaks to how stable the ordering is under
resampling — not to whether the ordering means anything substantively.

# Markov order: is a first-order model enough?

Every transition network here assumes the Markov property — that the next state
depends on the current one and nothing earlier. That assumption is testable.

```{r markov, eval=has_nestimate}
markov_order_test(walk_model)
```

The test disagrees with the model. AIC, BIC and the permutation LRT all prefer a
higher order, meaning yesterday carries information about tomorrow beyond today.
This does not invalidate the first-order network, but it bounds what it may
claim: a summary of one-step behaviour, not a full description of the process.

The natural follow-up is a higher-order network.

```{r hon, eval=has_nestimate}
build_hon(walk_model, max_order = 2)
```

HON's own promotion rule keeps every node at order 1 — it finds no individual
context that shifts the next-step distribution enough to justify splitting. Two
criteria disagreeing is a real result rather than a bug: the likelihood test sees
aggregate improvement across all sequences, while HON asks a stricter per-node
question. Report both.

# Comparing two periods

The tests so far describe one network. `permutation()` compares two, shuffling
the group labels to build a null distribution for every edge. The `steps` series
runs from April 2019 to March 2020; splitting at the start of November contrasts
the warmer months with the colder ones. The two spans are not equal in length —
roughly six and a half months against four.

Both models are built with the same fixed `breaks`, which matters more than it
looks: had each period been discretized on its own quantiles, the two networks
would sit on different cut points and any difference between them could be an
artefact of the thresholds rather than of behaviour.

```{r two-models, eval=has_nestimate}
cutoff <- as.Date("2019-11-01")

early <- ts_tna(
  subset(walkers, as.Date(day) < cutoff),
  value = "steps",
  id = "id",
  time = "day",
  discretization = "threshold",
  breaks = c(5000, 10000),
  labels = c("sedentary", "moderate", "active")
)

late <- ts_tna(
  subset(walkers, as.Date(day) >= cutoff),
  value = "steps",
  id = "id",
  time = "day",
  discretization = "threshold",
  breaks = c(5000, 10000),
  labels = c("sedentary", "moderate", "active")
)

comparison <- permutation(early, late, iter = 2000, seed = 2024)

comparison
```

```{r perm-table, eval=has_nestimate}
summary(comparison)
```

None of the nine edge comparisons reaches significance — the smallest p-value is
0.07. The estimates are not identical, but no difference survives the permutation
null. That is a more interesting result than a scattering of small p-values would
have been, because the two periods are *not* alike in how much people walked:

```{r composition, eval=has_nestimate}
state_distribution(early)

state_distribution(late)
```

Active days fall from 46.9% before November to 42.8% after it, and sedentary days
rise. So the seasonal shift shows up plainly in the *composition* of the states
while no comparison of the *transitions* between them reaches significance. The
estimated sedentary-to-moderate probability, for instance, is 0.446 before
November and 0.444 after, with a permutation p-value of 0.94.

Two cautions on how far that carries. Failing to reject nine tests is not the
same as showing the dynamics are identical, and the two models do not rest on
quite the same people — 140 participants contribute before November and 149
after, with 138 in both.

That distinction is still the reason to build a transition network at all. A
comparison of means or of state proportions finds the seasonal difference and
stops there; the transition model asks a further question about the same
population, and finds no evidence that the dynamics moved with it.

# More than two groups: one model per context

Building two models by hand is fine when there are two. When the thing that
separates the groups is already a column in the data, `group` builds one network
per level in a single call — and it scales past two, which the pairwise form
does not.

The `motivation` study records where each measurement was taken in
`task_context_type`, so the pleasure ratings can be split by context.

```{r grouped, eval=has_nestimate}
by_context <- ts_tna(
  motivation,
  series = "pleasure",
  group = "task_context_type",
  labels = c("low", "mid", "high")
)

by_context
```

Two things happen here that would be tedious to arrange by hand. The states are
cut from the *pooled* series before the split, so all four networks sit on one
set of thresholds and share a node set — the same discipline the fixed `breaks`
enforced in the previous section, applied automatically. And the result is a
Nestimate `netobject_group`, so the compatible grouped verbs take it directly:
`net_prune()`, `state_distribution()`, `net_centrality()`, `compare_model()`,
`permutation()`, `mosaic_plot()` and friends. Not every verb applies — the ones
that need a precision matrix or a clustering attribute reject a transition
model, and there is no grouped `plot()` method, so plot one group at a time.

`as.data.frame()` is the tidy view of how much data stands behind each network:

```{r grouped-index, eval=has_nestimate}
as.data.frame(by_context, what = "groups")
```

The sequence counts deserve a second look: `Home` has 1,324 observations but 832
sequences. That is not padding. A run is cut at every change of series *and* at
every change of group. The two cuts are not quite the same: a change of series
separates measurements that were never consecutive anyway, while a change of
context falls between two genuinely adjacent measurements, and the transition
across it is deliberately excluded from both networks — it belongs to neither
context. This study alternates
context from one measurement to the next — 4,871 observations form 2,632
contiguous runs — so each unbroken stretch within a context becomes its own
sequence. A pair of observations separated by an intervening visit to a different
context is never counted as a transition, because it never was one.

That rule costs nothing when a group is constant within a series, which is the
usual case for a condition or a cohort: each series contributes one sequence and
keeps its own ID.

```{r grouped-distribution, eval=has_nestimate}
state_distribution(by_context)
```

The contexts are not alike. Pleasure is low on 62.5% of measurements taken at
home and high on only 6.6% of them; personal time reverses that ordering — high
on 45.6%, low on 19.3%. Work sits between the two on all three states.

Because every network shares an alphabet, two of them can be set side by side and
picked out by name:

```{r grouped-compare, eval=has_nestimate}
compare_model(by_context, i = "Home", j = "Work")
```

Pruning applies group-wise and returns a grouped model, so the collection can be
carried through a workflow without being taken apart:

```{r grouped-prune, eval=has_nestimate}
net_prune(by_context, method = "threshold", threshold = 0.3)
```

One caution the table above states plainly rather than hides: `Other` holds three
observations and two sequences. It is a level of the column, so it gets a
network, but a network estimated from one transition is not a result. Read the
`sequences` and `observations` columns before reading the edges.

# Plotting

```{r plot, eval=has_nestimate && has_cograph, fig.width=9, fig.height=6, fig.alt="Transition network of daily step-count states, with sedentary, moderate, and active nodes joined by directed edges whose thickness reflects transition probability."}
plot(walk_model)
```

# Where the line falls

Everything descriptive in this vignette — centralities, state distributions,
entropy, pruning, per-participant splits, and the grouped models above — runs on
a model built from a single series. Everything inferential — bootstrap,
stability, the Markov-order test and the permutation test — needs many
sequences, because resampling and permutation both treat the sequence as the
unit of analysis.

Build with `id` from the start if you intend to test anything. Fix your `breaks`
before you compare anything, or pass `group` and let the pooled discretization
fix them for you. And when a comparison rests on few sequences, say so: the
`sequences` column exists to be read.
