---
title: "The Nestimate Compatibility Map"
author: "tsn package authors"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{The Nestimate Compatibility Map}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 110,
  out.width = "100%"
)
library(tsn)
has_nestimate <- requireNamespace("Nestimate", quietly = TRUE)
has_cograph <- requireNamespace("cograph", quietly = TRUE) &&
  requireNamespace("igraph", quietly = TRUE)
```

# One object, two packages

A `ts_tna()` model is not exported *to* Nestimate; it *is* a Nestimate
object. This vignette maps what that buys, verb by verb: which parts of
Nestimate's analysis surface accept a tsn transition model directly, what
question each verb answers, what each result draws, and which parts do not
apply and why. The companion vignette
`vignette("nestimate-workflow", package = "tsn")` runs a single analysis end
to end; this one is the reference you consult when deciding which verb to
reach for.

The running example is the model from that workflow: 151 participants' daily
step counts, cut at 5,000 and 10,000 steps so the states mean something
outside the sample.

```{r build, eval=has_nestimate}
library(tsn)
library(Nestimate)

data(steps)
walkers <- subset(steps, !is.na(steps))

walk_model <- ts_tna(
  walkers,
  value = "steps",
  id = "id",
  time = "day",
  discretization = "threshold",
  breaks = c(5000, 10000),
  labels = c("sedentary", "moderate", "active")
)

class(walk_model)
```

The class chain is the whole compatibility story. `ts_tna` is the tsn layer
(it remembers the source series and discretization), `netobject` is what
every Nestimate verb dispatches on, and `cograph_network` is what
`cograph::splot()` renders. Nestimate's own validator agrees:

```{r validate, eval=has_nestimate}
validate_netobject(walk_model)
```

And because the object is a `cograph_network`, the model itself is one
`plot()` away:

```{r plot-model, eval=has_nestimate && has_cograph, fig.width=6.5, fig.height=5, fig.alt="Directed transition network with sedentary, moderate, and active nodes; edge labels show transition probabilities, with the heaviest self-loop on the active state."}
plot(walk_model, type = "network")
```

The strong diagonal is the first substantive fact: every state prefers
itself, `active` most of all at 0.62.

# Reading the model as data

Both packages provide accessors, and they answer slightly different
questions. tsn's `as.data.frame()` returns every state pair including
self-transitions; `as.matrix()` returns the row-stochastic weight matrix.

```{r tsn-accessors, eval=has_nestimate}
head(as.data.frame(walk_model))
```

Nestimate's extractors return the same estimates in its own shapes:

```{r extract, eval=has_nestimate}
extract_transition_matrix(walk_model)

extract_initial_probs(walk_model)

extract_edges(walk_model)
```

One difference matters for transition networks: `extract_edges()` **excludes
self-transitions** — it returns the six between-state edges, sorted by
weight, and omits the diagonal. For this model the diagonal is where most of
the probability lives, so use `as.data.frame()` or
`extract_transition_matrix()` whenever persistence is part of the question,
and `extract_edges()` when only between-state movement is.

The initial-state probabilities say the panel starts active: 57.6% of
participants' first recorded day is an `active` day, 11.3% a `sedentary`
one.

# Where the process spends its time

`state_distribution()` and `state_frequencies()` count observations per
state — the marginal composition, before any dynamics.

```{r distribution, eval=has_nestimate}
state_distribution(walk_model)

state_frequencies(walk_model)
```

Active days make up 44.8% of the panel and sedentary days 14.0%. Three
displays draw the composition from different angles. The mosaic is the
proportion bar:

```{r mosaic, eval=has_nestimate, fig.width=6.5, fig.height=4, fig.alt="Mosaic plot of the step-count model's state composition, with active and moderate days forming most of the area and sedentary days a narrow band."}
mosaic_plot(walk_model)
```

`sequence_plot()` is the sequence carpet — one row per participant, one
coloured cell per day:

```{r carpet, eval=has_nestimate, fig.width=7, fig.height=5, fig.alt="Sequence heatmap with one row per participant and one coloured cell per day, showing long same-state spells as horizontal streaks of colour."}
sequence_plot(walk_model)
```

The streaks are persistence made visible: participants hold a state for
spells, not single days. `distribution_plot()` stacks the day-by-day state
mix of the whole panel:

```{r composition-time, eval=has_nestimate, fig.width=7, fig.height=4, fig.alt="Stacked area chart of the panel's state proportions across the study days, with bands for sedentary, moderate, and active days."}
distribution_plot(walk_model)
```

This is composition over calendar time — the view in which the seasonal
shift quantified in the workflow vignette (active days falling from 46.9% to
42.8% after November) lives. `plot_state_frequencies()` offers a fourth,
marimekko-style variant of the same information.

# Which states and edges carry the structure?

`net_centrality()` ranks states. Self-transitions are excluded by default —
for a transition network the diagonal would otherwise dominate every
strength measure.

```{r centrality, eval=has_nestimate}
walk_centrality <- net_centrality(walk_model)

walk_centrality
```

```{r centrality-plot, eval=has_nestimate, fig.width=7, fig.height=4, fig.alt="Centrality chart of the three states showing moderate with the highest in-strength and betweenness."}
plot(walk_centrality)
```

`moderate` is the hub: the highest in-strength (0.77) and the only state
with nonzero betweenness. That is the staircase structure — traffic between
`sedentary` and `active` routes through the middle.

`net_edge_betweenness()` asks the same question of edges:

```{r betweenness, eval=has_nestimate}
walk_betweenness <- net_edge_betweenness(walk_model)

walk_betweenness
```

```{r betweenness-plot, eval=has_nestimate, fig.width=6.5, fig.height=4.5, fig.alt="Edge-betweenness network view with the moderate-to-sedentary and active-to-moderate edges carrying the highest loads."}
plot(walk_betweenness)
```

The heaviest-loaded edges (betweenness 2) are `moderate → sedentary` and
`active → moderate`: the descending staircase. Movement down the activity
scale routes through the middle state.

# How the process moves

## Mean first passage times

`passage_time()` converts the transition matrix into expected travel times
between states (Kemeny & Snell, 1976).

```{r passage, eval=has_nestimate}
walk_passage <- passage_time(walk_model)

walk_passage
```

```{r passage-plot, eval=has_nestimate, fig.width=6.5, fig.height=4.5, fig.alt="Heatmap of mean first passage times between the three states, with long times into the sedentary state and short times into moderate and active."}
plot(walk_passage)
```

A `moderate` or `active` day is reached from anywhere within two to four
days on average; a `sedentary` day takes seven to ten. Low activity is the
state this panel visits reluctantly, not a basin it falls into. The
stationary distribution — 14% sedentary, 41% moderate, 45% active — is where
the chain settles regardless of its start.

## Persistence, return, and sojourn

`markov_stability()` summarizes each state's dynamics in one row: the
self-transition probability, the stationary share, the mean recurrence time,
and the mean sojourn (how long a visit lasts once entered).

```{r markov-stability, eval=has_nestimate}
walk_spells <- markov_stability(walk_model)

walk_spells
```

```{r markov-stability-plot, eval=has_nestimate, fig.width=7, fig.height=4, fig.alt="Bar panels of persistence, stationary probability, return time, and sojourn time for the three states."}
plot(walk_spells)
```

An active spell lasts 2.6 days on average and recurs every 2.2; a sedentary
spell lasts 1.6 days and recurs only every 7.1. Persistence and prevalence
line up here, but they are logically distinct columns — a rare state can
still be sticky.

## Is the chain well-behaved?

`chain_structure()` reports the Markov-chain diagnostics that other verbs
quietly assume: irreducibility (every state reachable from every other),
aperiodicity, and reversibility.

```{r chain, eval=has_nestimate}
summary(chain_structure(walk_model))
```

The chain is irreducible, aperiodic, and regular, so the stationary
distribution above is well-defined and unique. It is *not* reversible: the
forward process is statistically distinguishable from its time reversal,
which is itself a substantive fact about behavior change — the routes up and
the routes down are not mirror images.

# How predictable is the process?

`transition_entropy()` gives the point estimate; `entropy_bayes()` adds a
posterior credible interval by placing a Dirichlet prior on each row.

```{r entropy, eval=has_nestimate}
transition_entropy(walk_model)
```

```{r entropy-bayes, eval=has_nestimate}
set.seed(2026)
walk_entropy <- entropy_bayes(walk_model)

walk_entropy
```

```{r entropy-bayes-plot, eval=has_nestimate, fig.width=7, fig.height=4, fig.alt="Posterior distributions of the entropy rate, stationary entropy, and redundancy, each a narrow density around its point estimate."}
plot(walk_entropy)
```

The entropy rate is 1.346 bits against a ceiling of `log2(3) = 1.585` —
0.849 on the normalised scale — with a credible interval two hundredths
wide: with 30,000 transitions, the *uncertainty* about how unpredictable the
process is, is itself tiny. The redundancy of 0.098 bits is the gap between
the marginal entropy and the conditional one: knowing today's state removes
about 7% of tomorrow's uncertainty.

`entropy_network()` maps the same quantity onto the edges — each entry is an
edge's contribution to the row's unpredictability — and returns a network of
the same class, so it can be plotted or pruned like any other:

```{r entropy-network, eval=has_nestimate}
entropy_network(walk_model)
```

# Does history beyond one step matter?

The model assumes the next state depends only on the current one.
`markov_order_test()` (Anderson & Goodman, 1957) tests that assumption.

```{r markov-order, eval=has_nestimate}
set.seed(2026)
walk_order <- markov_order_test(walk_model)

walk_order
```

```{r markov-order-plot, eval=has_nestimate, fig.width=7, fig.height=4, fig.alt="Model-selection chart of log-likelihood, AIC, and BIC across Markov orders zero to three, all favouring orders above one."}
plot(walk_order)
```

BIC, AIC, and the permutation LRT all select order 3: yesterday carries
information about tomorrow beyond today. `path_dependence()` says how much:

```{r path-dependence, eval=has_nestimate}
walk_history <- path_dependence(walk_model)

walk_history
```

```{r path-dependence-plot, eval=has_nestimate, fig.width=7, fig.height=4, fig.alt="Bar chart of the KL divergence contributed by each two-step context, all values below 0.17 bits."}
plot(walk_history)
```

The weighted divergence is 0.035 bits: history shifts the next-step
distribution by a few hundredths of a bit and flips the modal prediction in
only two of nine contexts. This is the pair of numbers that reconciles a
decisively significant order test with the continued use of a first-order
model — detectable is not the same as large. The first-order network remains
a fair one-step summary; it is not the full process.

`path_counts()` is the raw material — every observed two-step path with its
count:

```{r path-counts, eval=has_nestimate}
path_counts(walk_model)
```

`active → active` alone is 27.8% of all transitions. And when the order test
demands more than one step, `build_hon()` estimates a higher-order network
under its own per-node promotion rule:

```{r hon, eval=has_nestimate}
hon <- build_hon(walk_model, max_order = 2)

pathways(hon)
```

HON promotes no node here — no single context shifts the next-step
distribution enough to justify splitting a state — which is why `pathways()`
comes back empty. Two criteria disagreeing is a result, not a bug: the
likelihood test sees aggregate improvement, HON asks a stricter per-node
question. Report both.

# Simplifying the model

`net_prune()` removes weak edges and returns the same class;
`net_pruning_details()` is the audit trail; `net_deprune()` undoes it.

```{r prune, eval=has_nestimate}
pruned <- net_prune(walk_model, method = "threshold", threshold = 0.3)

net_pruning_details(pruned)
```

Three edges fall: the two routes *into* sedentary and the direct
`sedentary → active` leap. Because pruning is recorded rather than
destructive, the original model is one call away:

```{r deprune, eval=has_nestimate}
net_deprune(pruned)
```

# How sure are we about the estimates?

Nestimate offers several answers resting on different machinery: sequence
resampling, node resampling, case dropping, sample splitting, and a
posterior.

## Sequence bootstrap

`bootstrap_network()` resamples the 151 sequences and re-estimates.

```{r bootstrap, eval=has_nestimate}
walk_boot <- bootstrap_network(walk_model, iter = 200, seed = 2026)

walk_boot
```

`summary()` is the tidy one-row-per-edge table:

```{r bootstrap-summary, eval=has_nestimate}
head(summary(walk_boot), 3)
```

All nine edges survive, with confidence intervals a few hundredths wide —
what 151 sequences buy.

## Bayesian edge certainty

`certainty()` reaches the same verdict from a Dirichlet posterior on each
row, with no resampling.

```{r certainty, eval=has_nestimate}
set.seed(2026)
certainty(walk_model)
```

All nine edges are certain. `summary()` returns the per-edge table with
posterior means, intervals, and flags.

## Whole-network statistics: the vertex bootstrap

`vertex_bootstrap()` implements the Snijders–Borgatti (1999) vertex
bootstrap for network-level statistics.

```{r vertex-boot, eval=has_nestimate}
vertex_bootstrap(walk_model, iter = 100, seed = 2026)
```

The mean edge weight carries a wide interval ([0.15, 0.35]) despite the huge
sample, because the vertex bootstrap resamples *nodes* and there are only
three — a deliberately conservative design for small node sets.

## Case dropping, in two flavours

`centrality_stability()` tracks the centrality *ordering* as sequences are
dropped; `casedrop_reliability()` tracks the edge weights themselves. Both
report the CS-coefficient of Epskamp, Borsboom & Fried (2018).

```{r stability, eval=has_nestimate}
set.seed(2026)
centrality_stability(walk_model, iter = 100)
```

```{r casedrop, eval=has_nestimate}
walk_reliability <- casedrop_reliability(walk_model, iter = 100,
                                         seed = 2026)

walk_reliability
```

```{r casedrop-plot, eval=has_nestimate, fig.width=7, fig.height=4, fig.alt="Case-dropping stability curve showing the correlation between full-sample and subsampled edge weights staying above 0.95 even when 90 percent of sequences are dropped."}
plot(walk_reliability)
```

Both CS-coefficients are 0.90 — the orderings and the weights survive
dropping 90% of the participants, far above the 0.7 bar for excellent. Even
at a 90% drop the mean edge-weight correlation with the full model is 0.957.

## Split-half reliability

```{r reliability, eval=has_nestimate}
set.seed(2026)
network_reliability(walk_model)
```

Randomly halving the panel 1,000 times yields two models that correlate at
0.99 with a mean absolute edge difference of 0.02. Every verb in this
section agrees: with 151 sequences, this network is estimated about as
precisely as a 3-state model can be.

# Comparing two models

Any two models on the same alphabet can be compared. The pair here splits
the panel at the start of November — the warm and cold seasons — with the
same fixed `breaks`, so both models sit on one alphabet.

```{r seasons, eval=has_nestimate}
cutoff <- as.Date("2019-11-01")

early <- ts_tna(
  subset(walkers, as.Date(day) < cutoff),
  value = "steps", id = "id", time = "day",
  discretization = "threshold", breaks = c(5000, 10000),
  labels = c("sedentary", "moderate", "active")
)

late <- ts_tna(
  subset(walkers, as.Date(day) >= cutoff),
  value = "steps", id = "id", time = "day",
  discretization = "threshold", breaks = c(5000, 10000),
  labels = c("sedentary", "moderate", "active")
)
```

## The difference itself

`subtract_networks()` returns the edge-by-edge difference as an object
cograph can draw:

```{r subtract, eval=has_nestimate}
season_difference <- subtract_networks(early, late)

season_difference
```

```{r subtract-plot, eval=has_nestimate && has_cograph, fig.width=6.5, fig.height=5, fig.alt="Difference network of the two seasonal models, edges coloured by the sign of the seasonal change and all magnitudes small."}
cograph::plot_difference(season_difference)
```

The largest seasonal gap on any edge is 0.042 (`active → active`).

## The permutation test

`permutation()` shuffles sequences between the two groups to build a null
distribution per edge.

```{r permutation, eval=has_nestimate}
season_test <- permutation(early, late, iter = 1000, seed = 2026)

summary(season_test)
```

No edge reaches significance; the smallest p-value is 0.07
(`active → moderate`). The seasons differ in *composition* (the workflow
vignette shows active days falling from 46.9% to 42.8%) while no evidence
places the difference in the *dynamics*.

## The metric panel

```{r compare, eval=has_nestimate}
season_panel <- compare_model(early, late)

season_panel
```

```{r compare-plot, eval=has_nestimate, fig.width=7, fig.height=4.5, fig.alt="Comparison panel chart of the two seasonal models across weight-deviation, correlation, and similarity metrics, all indicating near-identity."}
plot(season_panel)
```

A Pearson correlation of 0.995 between the two matrices: the seasons share
their dynamics almost entirely.

## The instructive disagreement

`bayes_compare()` asks whether each edge difference is *credibly nonzero*:

```{r bayes-compare, eval=has_nestimate}
set.seed(2026)
season_bayes <- bayes_compare(early, late)

season_bayes
```

```{r bayes-compare-plot, eval=has_nestimate, fig.width=7, fig.height=4.5, fig.alt="Posterior difference intervals for the nine edges, four of which exclude zero while all magnitudes stay below 0.05."}
plot(season_bayes)
```

Four of nine edges are credibly different, while the permutation test above
found none significant. Both are right, about different questions. With
roughly 30,000 transitions, differences of two to four hundredths are
estimated precisely enough for the posterior to exclude zero (the
credible-difference threshold here is 0.01) — but the sequence-level
permutation null, which respects the fact that days are nested in people,
absorbs them. Precision detects that the difference exists; it does not make
0.04 large. Report the magnitude next to whichever test is used.

`vertex_compare()` runs the node-resampling comparison at the level of
whole-network statistics and, consistent with all of the above, finds
nothing (all p ≥ 0.78):

```{r vertex-compare, eval=has_nestimate}
vertex_compare(early, late, iter = 200, seed = 2026)
```

# Link prediction

`predict_links()` scores unobserved dyads by six structural predictors:

```{r predict, eval=has_nestimate}
predict_links(walk_model)
```

It runs, but a three-state transition network is fully connected between
states, which leaves link prediction nothing to rank; it earns its place on
larger, pruned alphabets, where a missing edge is a real hypothesis.

# Grouped models

When the comparison variable is a column of the data,
`ts_tna(group = ...)` builds one network per level and returns a Nestimate
`netobject_group`. The grouped verbs verified against it are
`net_centrality()`, `state_distribution()`, `transition_entropy()`,
`passage_time()`, `certainty()`, `net_prune()`, `compare_model()`,
`permutation()`, `bootstrap_network()`, `markov_order_test()`,
`centrality_stability()`, `casedrop_reliability()`, and `mosaic_plot()`;
`summary()` and `as.data.frame()` return tidy per-group tables, and
`plot()` draws one selected group. `net_edge_betweenness()` also runs, but
its result drops back to a plain `netobject_group`. The dedicated vignette
`vignette("group-models", package = "tsn")` works through all of it.

# What does not apply, and why

Three families of Nestimate verbs reject a transition model, each for a
principled reason rather than a missing feature:

* **Sequence-data verbs.** `frequencies()`, `mark_first_state()`,
  `mark_terminal_state()`, `entropy_trajectory()`, `mosaic_analysis()`,
  `plot_mosaic()`, and `as_htna()` operate on a raw wide data frame of
  sequences, not on an estimated model. They belong *before* model
  construction in a workflow, applied to sequence data you hold yourself.
* **Psychometric-network verbs.** `predictability()` and the
  `glasso`/`pcor`/`cor` estimator family need a precision or correlation
  structure estimated from multivariate data. A transition matrix is a
  conditional-probability object, not a conditional-independence one; the
  quantities these verbs compute are undefined for it.
* **Verbs that need extra structure.** `cluster_network()` and
  `build_clusters()` require a chosen number of clusters and are aimed at
  multi-sequence typologies; `wtna()` needs one-hot action data;
  `sequence_compare()` wants sequence data plus a grouping vector. And
  `as_tna()` is unnecessary by design — a tsn model needs no conversion,
  which is the point of this vignette.

# The compatibility map

| Question | Verb | Tidy access | Plot | Notes |
|---|---|---|---|---|
| What are the estimates? | `as.data.frame()`, `as.matrix()`, `extract_transition_matrix()`, `extract_initial_probs()` | direct | via `plot()` on the model | tsn accessors include the diagonal |
| Between-state edges only | `extract_edges()` | direct | — | **excludes self-transitions** |
| Marginal composition | `state_distribution()`, `state_frequencies()` | data frame | `mosaic_plot()`, `plot_state_frequencies()` | |
| Composition over time / carpet | `distribution_plot()`, `sequence_plot()` | — | draws directly | stacked area; sequence heatmap |
| State/edge importance | `net_centrality()`, `net_edge_betweenness()` | data frame / network | `plot()` on both | loops excluded from centrality by default |
| Travel times between states | `passage_time()` | print | `plot()` | Kemeny & Snell (1976) |
| Persistence, recurrence, sojourn | `markov_stability()` | `$stability` table | `plot()` | |
| Chain diagnostics | `chain_structure()` | `summary()` | `plot()` heatmap | irreducibility, reversibility |
| Predictability | `transition_entropy()`, `entropy_bayes()`, `entropy_network()` | print / `summary()` | `plot()` on both entropy results | rate in bits; ceiling `log2(n_states)` |
| History beyond one step | `markov_order_test()`, `path_dependence()`, `path_counts()`, `build_hon()` + `pathways()` | print / data frame | `plot()` on test and dependence | test the size, not only the p |
| Simplification | `net_prune()`, `net_pruning_details()`, `net_deprune()` | same class back | via model `plot()` | prune after testing |
| Edge uncertainty | `bootstrap_network()`, `certainty()` | `summary()` | — | resampling vs Bayesian |
| Network-level uncertainty | `vertex_bootstrap()` | data frame | `plot()` | Snijders & Borgatti (1999) |
| Sample-size reliability | `centrality_stability()`, `casedrop_reliability()`, `network_reliability()` | print | `plot()` on all three | CS-coefficient bar: 0.5 / 0.7 |
| Two-model comparison | `permutation()`, `compare_model()`, `subtract_networks()`, `bayes_compare()`, `vertex_compare()` | `summary()` / data frame | `plot()`; `cograph::plot_difference()` | report magnitudes with tests |
| Link prediction | `predict_links()` | print | — | needs a sparse alphabet to be informative |
| Many-group comparison | `ts_tna(group=)` + grouped verbs | `as.data.frame()` | `plot(x, group=)`, `mosaic_plot()` | see `vignette("group-models")` |
| Rendering | `plot()`, `cograph::splot()` | — | — | requires cograph |

# References

Anderson, T. W., & Goodman, L. A. (1957). Statistical inference about
Markov chains. *The Annals of Mathematical Statistics*, 28(1), 89–110.

Epskamp, S., Borsboom, D., & Fried, E. I. (2018). Estimating psychological
networks and their accuracy: A tutorial paper. *Behavior Research Methods*,
50(1), 195–212.

Kemeny, J. G., & Snell, J. L. (1976). *Finite Markov Chains*. Springer.

Snijders, T. A. B., & Borgatti, S. P. (1999). Non-parametric standard
errors and tests for network statistics. *Connections*, 22(2), 161–170.

Xu, J., Wickramarathne, T. L., & Chawla, N. V. (2016). Representing
higher-order dependencies in networks. *Science Advances*, 2(5), e1600028.
