---
title: "The tsn Function Set Applied to One Intensive Longitudinal Series"
author: "tsn package authors"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{The tsn Function Set Applied to One Intensive Longitudinal Series}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 150,
  out.width = "100%"
)
library(tsn)
# cograph's centrality node sizing (the default for network plots) calls
# igraph, which cograph only Suggests, so require both here.
has_cograph <- requireNamespace("cograph", quietly = TRUE) &&
  requireNamespace("igraph", quietly = TRUE)
has_nestimate <- requireNamespace("Nestimate", quietly = TRUE)
```

# Time Series as Networks

A time series can be transformed into a network representation by mapping its
temporal observations or patterns onto a set of nodes and defining edges
according to a specific relationship between them. This transformation provides
an alternative way of analyzing temporal dynamics, allowing techniques from
network science to be applied to data that are originally represented as
sequences of numerical observations. Different network-construction methods
emphasize different aspects of the underlying time series. In **state-transition
networks**, the continuous or discrete values of a time series are first
discretized into a finite number of states, with nodes representing these states
and edges representing transitions between successive observations. In
**distance-based networks**, observations, windows, or subsequences are
represented as nodes and connected when their distance according to a selected
metric, such as Euclidean distance or dynamic time warping, satisfies a given
threshold or weighting rule. **Similarity networks** use measures such as
correlation, mutual information, or other similarity functions to connect
observations or temporal segments that exhibit similar behavior. A related
approach is based on **proximity**, where nodes are connected according to how
close their values or local temporal patterns are, producing a network that
captures the geometric structure of the data. **Visibility graphs** provide a
different type of transformation: each observation becomes a node, and two nodes
are connected when the corresponding points in the time series satisfy a
geometric visibility criterion. Natural visibility graphs and horizontal
visibility graphs are prominent examples of this approach. **Recurrence
networks** connect observations when the underlying system returns to
sufficiently similar states, thereby capturing recurrent structures and aspects
of the system's phase-space dynamics. Another approach is based on **ordinal
patterns**, in which nodes represent the relative ordering of consecutive
observations rather than their absolute values, making the resulting network
particularly useful for characterizing temporal fluctuations and nonlinear
dynamics. Together, these approaches provide a broad framework for representing
time series as networks, with each method highlighting different properties of
the data, including transitions, distances, similarities, proximity, visibility,
recurrence, and ordinal structure. The resulting networks can then be studied
using measures such as degree distributions, centrality, clustering, motifs,
community structure, and connectivity to identify patterns and changes in the
dynamics of the original time series.

The tsn package reaches all of these constructions through two arguments of one
constructor. `method` selects the family and `unit` selects what a node
represents, so a single call expresses each approach named above.
State-transition networks come from the `ts_tna()` family, which discretizes the
series and passes the state sequence to Nestimate. Setting `unit = "state"` in
`tsn()` aggregates the edges of a visibility or distance network up to the state
level under the rule in `aggregation`, which produces a state-level graph of the
selected construction. Distance-based networks come from `tsn(x, "distance")`, whose
`distance` argument covers Euclidean, Manhattan, dynamic time warping, and
further metrics. Similarity networks are the same family with `distance` set to
`"correlation"`, `"spearman"`, `"nmi"`, or `"cosine"`. Proximity enters through
the `connect` argument, which retains edges by nearest neighbours, by a distance
threshold, by a percentile of the shortest distances, or by a Gaussian kernel.
Visibility graphs come from `tsn(x, "nvg")` and `tsn(x, "hvg")`, and from the
`vg()` verb. Recurrence networks are distance networks over individual time
points under a threshold rule, `tsn(x, "distance", unit = "time", connect =
"threshold")`. Ordinal-pattern networks come from `tsn(x, "ordinal")`, which
applies the Bandt-Pompe embedding and uses the resulting permutation patterns as
nodes.

This vignette applies the nine exported functions to one measured series,
following the construction pipeline from discretization through the trend
classifier, the visibility and distance families, the four transition
constructors, and the per-series split. The functions and the objects they
return:

| Function | Returned object | Nodes | Edge weight |
|---|---|---|---|
| `discretize()` | `tsn_states` table | none | none |
| `trend()` | `tsn_trend` table | none | none |
| `vg()` | `tsn` network | time points | 1 for a visibility relation |
| `tsn()` | `tsn` network | series, windows, time points, or states | similarity derived from a distance |
| `ts_ftna()` | `ts_tna` network | states | transition count |
| `ts_tna()` | `ts_tna` network | states | transition probability |
| `ts_cna()` | `ts_tna` network | states | co-occurrence count |
| `ts_atna()` | `ts_tna` network | states | decay-weighted transition |
| `series_networks()` | list of `ts_tna` networks | states | as in the pooled model |

Network figures are drawn by cograph. The four transition constructors and
`series_networks()` require Nestimate.

# The series

The `motivation` data set holds 4,871 repeated measurements of motivation,
appraisal, context, and mood, recorded from 2018-09-29 through 2021-07-12 with
up to seven measurements per day. The `pleasure` column is an integer appraisal
measure and is the single variable used throughout this vignette.

```{r data}
data("motivation", package = "tsn")

pleasure <- head(motivation$pleasure, 300)

length(pleasure)
summary(pleasure)
```

The first 300 observations are used. The restriction keeps the visibility graph
readable, since a graph on 4,871 nodes cannot be inspected in a rendered
document. No values are altered and the measurement order is preserved.

The distribution runs from 2 to 44 with a median of 21 and a mean of 21.4. The
closeness of the median and the mean indicates an approximately symmetric
distribution over this segment.

```{r raw-plot}
plot(pleasure, type = "l")
```

The horizontal axis is observation order. The series oscillates over the full
range of the scale without a visible drift, and the amplitude of the
oscillation varies across the segment.

# `discretize()`

`discretize()` assigns every observation to one of a fixed number of states and
returns a tidy table with one row per observation. The function exposes the
package's state discretizers as a single verb. `tsn()` routes its own state
aggregation through the same engine, so `discretize(x, method = "quantile")`
and `tsn(x, "quantile")` produce identical state assignments.

The default method is `"quantile"` with three states, which places the interior
cut points at the empirical tertiles of the pooled values. Quantile
discretization is the appropriate default when the states are meant to carry
equal weight in the analysis, because it targets equal counts rather than equal
value ranges.

```{r discretize}
states <- discretize(pleasure, labels = c("Low", "Middle", "High"))

states
```

The returned object has class `tsn_states` and carries the columns `id`,
`time`, `value`, `state`, and `probability`. The `id` column names the series,
which is `series_1` for a bare numeric vector. The `time` column holds the
observation index. The `probability` column records the confidence of the state
assignment and equals 1 throughout, because quantile discretization is
deterministic: a value falls on one side of a cut point or the other.

```{r discretize-summary}
summary(states)
```

`summary()` on a `tsn_states` object reports the count, the proportion, and the
mean value of each state. The three states hold 104, 104, and 92 observations.
Quantile discretization targets 100 observations per state here, and the
departure from that target follows from ties in the integer ratings. All
observations sharing a value must receive the same state, so a value sitting on
a cut point moves as a block and the counts cannot be equalized exactly.

The state means of 10.8, 21.6, and 33.1 increase across Low, Middle, and High,
so the state order matches the value order and the labels are correctly
assigned.

```{r discretize-plot}
plot(states, "ribbon")
```

`plot()` on a `tsn_states` object draws the original series together with its
state assignment. The `"ribbon"` type places the states in a strip beneath the
series. The default `"overlay"` type shades the panel itself, which suits a
series that holds each state over long stretches. This series changes state at
almost every observation, so panel shading would obscure the measurements and
the strip is the readable choice.

The strip shows how long the series stays within one state before crossing a cut
point. That duration is the property the transition networks in Section 7
quantify.

# `trend()`

`trend()` computes a rolling trend metric at every observation and classifies
each point into one of six classes: `"Ascending"`, `"Descending"`, `"Flat"`,
`"Turbulent"`, `"Missing Data"`, and `"Initial"`. The classification proceeds
in three steps. The function first fits a slope within a moving window. It then
applies a flat band of half-width `epsilon` around zero, so that slopes above
`+epsilon` are ascending, slopes below `-epsilon` are descending, and slopes
inside the band are flat. It finally overrides the label with `"Turbulent"`
where a rolling volatility measure exceeds a threshold.

The default slope estimator is the Theil-Sen estimator, which takes the median
of the pairwise slopes within the window. The Theil-Sen slope is resistant to
outlying observations, a property that matters for intensive longitudinal
measures where isolated extreme ratings are common.

The `window` argument sets the width of the moving window. When `window` is
left at `NULL`, the function uses `max(3, min(n, round(n / 10)))`, which gives
30 observations for this series.

```{r trend}
directions <- trend(pleasure)

directions
```

The returned object has class `tsn_trend` with the columns `id`, `time`,
`value`, `metric`, and `state`. The `metric` column holds the rolling slope and
`state` holds its classification.

```{r trend-summary}
summary(directions)
```

Descending covers 106 observations and Turbulent 77, so falling and volatile
segments together account for 61% of the series. Ascending covers 52
observations. The excess of descending over ascending points reflects the
asymmetry of the oscillation rather than a net decline, since the series begins
and ends at comparable levels: declines in this segment are spread over more
observations than the recoveries.

The 29 `"Initial"` observations are the first 14 and the last 15 points. The
default alignment is `"center"`, so the metric is assigned to the midpoint of
each window and a window of width 30 has no complete span at either end of the
series. Those points receive a missing metric and the `"Initial"` label. The
`"Missing Data"` class does not appear because the `motivation` data set
contains no missing values.

```{r trend-plot}
plot(directions)
```

The default plot type is `"points"`, which colours each observation by its
trend class.

```{r trend-panels}
plot(directions, "panels")
```

The `"panels"` type places the series above the rolling metric that produced
the classification. The lower panel shows the metric crossing the flat band,
and reading the two panels together relates each labelled segment to the slope
value that generated the label.

# `vg()`

`vg()` constructs a visibility graph and returns a network object in which each
observation is a node. Two visibility rules are available. The natural
visibility rule connects two observations when the straight line joining the
tops of their values passes above every value between them. The horizontal
visibility rule connects two observations when every intervening value is
smaller than both endpoints.

`vg()` is a wrapper around `tsn()` with `method = "visibility"`, so `vg(x)` is
equivalent to `tsn(x, "nvg")` and `vg(x, "horizontal")` is equivalent to
`tsn(x, "hvg")`.

```{r vg-natural}
natural <- vg(pleasure)

summary(natural)
```

`summary()` on a `tsn` object returns one row describing the graph: the
construction method, the node unit, the counts of nodes, evaluated dyads, and
retained edges, the density, the range of edge weights, and the direction
setting. The natural visibility graph on 300 observations has 825 edges and a
density of 0.018.

```{r vg-horizontal}
horizontal <- vg(pleasure, "horizontal")

summary(horizontal)
```

The horizontal visibility graph on the same series has 560 edges and a density
of 0.012. The horizontal rule is the stricter of the two: any pair it connects
also satisfies the natural rule, so the horizontal graph is a subgraph of the
natural graph and its edge count is bounded above by 825. Both graphs carry
weights of 1, because a visibility relation either holds between two
observations or does not, and there is no intermediate strength to record.

`plot()` on a `tsn` object renders the network through cograph, and the
arguments in `...` are passed to `cograph::splot()`. A graph on 300 nodes
exceeds what a printed figure can resolve, so the figure below applies the same
construction to the first 40 observations, where individual visibility relations
stay distinguishable.

```{r vg-network, eval=has_cograph}
short <- head(motivation$pleasure, 40)

plot(vg(short, "horizontal"), layout = "spring", labels = FALSE,
     scale_nodes_by = "degree", node_size_range = c(2, 9),
     node_fill = "#4E79A7", edge_label_style = "none",
     edge_color = "#B9C2CC")
```

The graph combines a chain along the temporal order with longer edges that span
the intervening observations. Consecutive observations are always mutually
visible, which supplies the chain. Node size encodes degree in the figure, so
the observations that see furthest appear as the largest nodes.

Node degree in the 300-observation horizontal graph ranges from 1 to 13 with a
mean of 3.7. High values accumulate the long edges, because a peak remains
visible across the smaller observations on either side of it. The rank
correlation between the measured value and the node degree is 0.66 in this
segment, and the mean degree of the highest decile of values is 6.8 against 2.2
for the lowest decile.

```{r vg-series}
plot(horizontal, "series")
```

The `"series"` type draws the observations behind the network. The view is
specific to tsn, because it displays the original measurements rather than a
generic graph.

# `tsn()`

`tsn()` is the core constructor and reaches both the distance and the
visibility families. The `method` argument selects the family, `unit` selects
what a node represents, and the remaining arguments configure the distance
measure and the rule that converts distances into edges. The default is a
natural visibility graph over time points, so `tsn(pleasure)` returns the graph
already built by `vg(pleasure)` in Section 5.

A distance network over sliding windows treats each window as a node. `tsn()`
cuts the series into windows of width `window`, advances the start by `step`
observations, computes a distance between every pair of windows, and applies
the connection rule named in `connect`. The construction answers a different
question from the visibility graph: it describes which segments of the series
have similar shape, rather than which individual observations see one another.

```{r tsn-distance}
windows <- tsn(pleasure, "distance", step = 10)

summary(windows)
```

The window width defaults to 30 observations. With `step = 10` the 300
observations yield 28 overlapping windows, so the network has 28 nodes. The
default connection rule is `"full"`, which retains every evaluated pair. The
result is a complete graph on 28 nodes with 378 edges and a density of 1.

A complete graph places all of the information in the edge weights. The
`weight` column reports similarity rather than distance, under the default rule
`1 / (1 + d)`, so larger weights correspond to windows of more similar shape.
The weights range from 0.0095 to 0.0243, which corresponds to Euclidean
distances of roughly 40 to 104 between windows.

```{r tsn-nearest}
sparse <- tsn(pleasure, "distance", step = 10, connect = "nearest", neighbors = 2)

summary(sparse)
```

`connect = "nearest"` retains only the edges from each node to its closest
neighbours. With `neighbors = 2` the same 378 evaluated dyads reduce to 37
retained edges and the density falls to 0.098. The `dyads` and `edges` columns
differ for the first time here, because the connection rule discards evaluated
pairs. Sparsification of this kind is the standard preparation for community
detection and path-based measures, which are uninformative on a complete graph.

```{r tsn-network, eval=has_cograph}
plot(sparse, layout = "spring", labels = FALSE,
     scale_nodes_by = "degree", node_size_range = c(3, 10),
     node_fill = "#F28E2B", edge_label_style = "none",
     edge_color = "#B9C2CC")
```

The retained edges join windows of similar shape. Windows adjacent in time
share 20 of their 30 observations at this step size, which makes an edge
between neighbours a reasonable expectation. Two of the 37 retained edges join
adjacent windows, and the remaining 35 span gaps of up to 19 window positions.
Overlap in observations therefore does not determine Euclidean proximity for
this series. An oscillating series carries heavily overlapping windows into
different parts of the range, while windows separated by many positions can
align closely in shape. The distance network reports recurrence of shape rather
than contiguity in time, and the two properties come apart here.

# Transition networks

The four transition constructors take a sequence of states and return a network
in which nodes are states and edges are relations between them. Each function
discretizes its input when given raw values, and uses the assignments unchanged
when given a `discretize()` result. All four calls below receive `states`, so
the four networks describe the same Low, Middle, and High sequence and differ
only in the quantity placed on the edges.

## `ts_ftna()`

`ts_ftna()` counts adjacent state transitions and returns a directed network
whose edge weights are those counts. The entry in row *i* and column *j* is the
number of times state *i* is immediately followed by state *j*.

```{r ts-ftna, eval=has_nestimate}
counts <- ts_ftna(states)

counts
```

The printed weight matrix sums to 299, which is the number of adjacent pairs in
a sequence of 300 observations. Every transition is counted once, so the total
is fixed by the sequence length and carries no information about the dynamics.

The diagonal entries of 43, 44, and 36 are the largest in their respective
rows. The sequence therefore repeats its current state more often than it moves
to either alternative. Raw counts confound this persistence with the state
frequencies, since a state visited often supplies more opportunities for any
transition out of it. `ts_tna()` removes that confound.

```{r ts-ftna-plot, eval=has_nestimate && has_cograph, fig.width=9}
plot(counts, ribbon = TRUE, overlay = "none")
```

The default plot type is `"combined"`, which places the state sequence on the
left and the network on the right. Reading the two together relates a thick
edge in the network to the stretches of the sequence that produced it.

## `ts_tna()`

`ts_tna()` divides each row of the transition counts by its row total and
returns a directed network whose edge weights are conditional probabilities.
The entry in row *i* and column *j* estimates the probability of occupying
state *j* at the next observation given state *i* at the current one. Each row
of the printed matrix sums to one.

```{r ts-tna, eval=has_nestimate}
probabilities <- ts_tna(states)

probabilities
```

Row normalization makes the rows comparable across states of different
frequency, which is the reason to prefer probabilities over counts when
comparing the outgoing behaviour of states.

The diagonal holds the largest probability in every row: 0.417 for Low, 0.423
for Middle, and 0.391 for High. Each value exceeds the corresponding marginal
state proportion of 0.347, 0.347, and 0.307 reported by `summary(states)`.
Under independence between consecutive observations, the probability of moving
to state *j* would equal the marginal proportion of *j* regardless of the
current state. The three diagonal entries stand above their marginals, which
provides evidence that the series persists in its current state beyond what the
state frequencies alone would produce.

```{r ts-tna-plot, eval=has_nestimate && has_cograph, fig.width=9}
plot(probabilities, ribbon = TRUE, overlay = "none")
```

## `ts_cna()`

`ts_cna()` counts how often two states occur together within the same sequence
and returns an undirected network. The rule ignores order, so the weight matrix
is symmetric.

```{r ts-cna, eval=has_nestimate}
cooccurrence <- ts_cna(states)

cooccurrence
```

With a single sequence, the co-occurrence counts follow from the state counts
alone. Each unordered pair of positions holding states *i* and *j* contributes
one count, so an off-diagonal entry is the product of two state counts and a
diagonal entry is the number of unordered pairs within one state. The Low to
Middle entry of 10,816 equals 104 × 104. The Low diagonal of 5,356 equals
104 × 103 / 2, and the High diagonal of 4,186 equals 92 × 91 / 2.

Co-occurrence measured over one sequence therefore carries no information about
ordering, and its weights are a deterministic function of the marginal state
distribution. The measure becomes informative when several sequences are
compared, because the states that appear together within the same sequence then
vary across units.

```{r ts-cna-plot, eval=has_nestimate && has_cograph, fig.width=9}
plot(cooccurrence, ribbon = TRUE, overlay = "none")
```

## `ts_atna()`

`ts_atna()` builds an attention network, in which a transition is weighted by a
decay function of the gap between the two positions rather than counted
equally. A state that follows immediately contributes more weight than one that
follows after a delay. The result is directed.

```{r ts-atna, eval=has_nestimate}
attention <- ts_atna(states)

attention
```

The weights range from 15.06 to 25.58 with a mean of 19.30. The values are
accumulated attention, so they are neither integers nor probabilities, and the
rows do not sum to one. Comparison across the four networks should therefore
use the ordering of the entries rather than their magnitudes.

The diagonal again holds the largest value in each row, which agrees with the
persistence already visible in the counts and the probabilities. Agreement
across three weighting schemes indicates that the persistence is a property of
the sequence rather than an artefact of one construction.

```{r ts-atna-plot, eval=has_nestimate && has_cograph, fig.width=9}
plot(attention, ribbon = TRUE, overlay = "none")
```

# `series_networks()`

`series_networks()` splits a pooled transition model into one model per source
series and returns a named list. Each element is rebuilt with the network type
of the pooled model and with the shared state alphabet, so the per-series
networks carry the same nodes in the same order. The common node set is what
makes the elements directly comparable with one another and with the pooled
model.

```{r series-networks, eval=has_nestimate}
per_series <- series_networks(probabilities)

length(per_series)
names(per_series)
```

This vignette supplies one series, so the split returns a list of length one
named `series_1`, and that element reproduces the pooled model. The function
applies to the case where several series are pooled: the pooled model then
describes the group, each element describes one individual, and the shared
alphabet supports comparison of individual networks against the group network.

# Summary

| Function | Result on this series |
|---|---|
| `discretize()` | 104 Low, 104 Middle, and 92 High observations |
| `trend()` | 106 descending, 77 turbulent, 52 ascending, 36 flat, 29 initial |
| `vg()` | 825 natural and 560 horizontal visibility edges on 300 nodes |
| `tsn()` | 28 window nodes, complete at density 1, or 37 edges under nearest-neighbour |
| `ts_ftna()` | 299 transitions with the diagonal largest in each row |
| `ts_tna()` | Diagonal probabilities of 0.417, 0.423, and 0.391 |
| `ts_cna()` | Symmetric counts determined by the marginal state distribution |
| `ts_atna()` | Decay-weighted transitions from 15.06 to 25.58 |
| `series_networks()` | One model, `series_1`, for the one supplied series |

The nine functions describe different properties of the same 300 observations
in their original order. Each result is a description of one series. A single
sequence does not supply the independent replications that bootstrap and
permutation procedures assume, so the estimates reported here carry no standard
errors and support no inference about the population from which the series was
drawn. Analyses of that kind require several series. Every verb used here
accepts them through the same arguments, with additional series supplied as
additional `id` values in long format or as additional columns in wide format.
