---
title: "Every tsn Function with Pleasure"
author: "tsn package authors"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Every tsn Function with Pleasure}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 9,
  fig.height = 5,
  dpi = 120
)
library(tsn)
# cograph's centrality node sizing (the default for network plots) calls
# igraph, which cograph only Suggests -- so require both here.
has_cograph <- requireNamespace("cograph", quietly = TRUE) &&
  requireNamespace("igraph", quietly = TRUE)
has_nestimate <- requireNamespace("Nestimate", quietly = TRUE)
```

This tutorial uses one variable:

```r
motivation$pleasure
```

There is no context grouping, balancing, pseudo-time construction, reordering,
or transformation. The same consecutive series is passed to all nine exported
functions.

The complete vector has 4,871 observations. A 300-observation prefix is used
because a 4,871-node network is not readable in an HTML tutorial and the
co-occurrence model becomes unnecessarily expensive. The selection is simply
`motivation$pleasure[1:300]`; no values are changed.

Network plots are drawn exclusively by **cograph**. The four transition
constructors and `series_networks()` require **Nestimate**.

# Data

```{r data}
data("motivation", package = "tsn")

pleasure_full <- motivation$pleasure
pleasure <- pleasure_full[seq_len(300L)]

class(pleasure_full)
length(pleasure_full)
head(pleasure_full)
summary(pleasure_full)

class(pleasure)
length(pleasure)
head(pleasure)
summary(pleasure)

stopifnot(
  is.integer(pleasure),
  length(pleasure) == 300L,
  !anyNA(pleasure),
  identical(pleasure, motivation$pleasure[seq_len(300L)])
)
```

The x-axis below is observation order.

```{r raw-plot}
plot(
  pleasure,
  type = "l",
  xlab = "Observation",
  ylab = "Pleasure",
  main = "The 300 pleasure observations used throughout"
)
```

# `discretize()`

`discretize()` converts the numeric values into states. Here, quantile
boundaries create Low, Middle, and High pleasure states. Tied ratings mean the
three state counts need not be identical.

```{r discretize}
pleasure_states <- discretize(
  pleasure,
  method = "quantile",
  n_states = 3L,
  labels = c("Low", "Middle", "High")
)

class(pleasure_states)
head(pleasure_states)
table(pleasure_states$state)
attr(pleasure_states, "breaks")
```

This plot combines the original series with state shading and state-boundary
guides.

```{r discretize-combined}
plot(
  pleasure_states,
  type = "overlay",
  overlay = "vertical",
  lines = "horizontal",
  points = FALSE,
  xlab = "Observation",
  ylab = "Pleasure"
)
```

# `trend()`

`trend()` classifies the local direction of the same series. A centered window
of 30 means 30 observations, not 30 days.

```{r trend}
pleasure_trend <- trend(
  pleasure,
  window = 30L,
  method = "slope",
  slope = "robust",
  align = "center"
)

class(pleasure_trend)
head(pleasure_trend)
table(pleasure_trend$state)
summary(pleasure_trend)
```

The panels view combines pleasure with the rolling slope used for
classification.

```{r trend-combined}
plot(
  pleasure_trend,
  type = "panels",
  xlab = "Observation",
  ylab = "Pleasure"
)
```

# `vg()`

`vg()` constructs a visibility graph. The horizontal visibility rule connects
two observations when every value between them is lower than both endpoints.

```{r vg}
pleasure_hvg <- vg(pleasure, type = "horizontal")

class(pleasure_hvg)
summary(pleasure_hvg)
head(as.data.frame(pleasure_hvg))
```

The network plot below is cograph output. There is no second or fallback
network renderer.

```{r vg-network, eval=has_cograph}
plot(
  pleasure_hvg,
  layout = "spring",
  labels = FALSE,
  node_fill = "#4E79A7"
)
```

The source-series view is local to tsn and displays the data behind the
network.

```{r vg-series}
plot(
  pleasure_hvg,
  type = "series",
  xlab = "Observation",
  ylab = "Pleasure"
)
```

# `tsn()`

`tsn()` is the general constructor. Here it makes overlapping
30-observation windows, advances ten observations at a time, measures
correlation distance, and connects each window to its two nearest neighbors.

```{r tsn}
pleasure_windows <- tsn(
  pleasure,
  method = "distance",
  unit = "window",
  window = 30L,
  step = 10L,
  distance = "correlation",
  connect = "nearest",
  neighbors = 2L,
  similarity = "normalized_inverse"
)

class(pleasure_windows)
summary(pleasure_windows)

head(as.data.frame(pleasure_windows))
head(as.data.frame(pleasure_windows, connected = TRUE))
```

`as.data.frame()` contains every evaluated dyad, so the second table filters
to retained edges. Larger connected weights mean more similar window shapes.

```{r tsn-network, eval=has_cograph}
plot(
  pleasure_windows,
  layout = "spring",
  labels = TRUE,
  node_fill = "#F28E2B"
)
```

## Common network accessors

```{r network-accessors}
head(as.data.frame(pleasure_windows))
head(as.data.frame(pleasure_windows, what = "series"))
dim(as.matrix(pleasure_windows))
print(pleasure_windows)
summary(pleasure_windows)
```

# Transition-network functions

All four transition functions reuse `pleasure_states`. They therefore operate
on exactly the same Low--Middle--High sequence.

## `ts_ftna()`: frequency transition network

`ts_ftna()` reports raw adjacent-transition counts. A 300-observation sequence
contains 299 transitions.

```{r ts-ftna, eval=has_nestimate}
pleasure_ftna <- ts_ftna(pleasure_states)

class(pleasure_ftna)
pleasure_ftna$weights
sum(pleasure_ftna$weights)
```

### Combined plot

The left panel is the pleasure state sequence; the right panel is the cograph
frequency network.

```{r ts-ftna-combined, eval=has_nestimate && has_cograph, fig.width=11, fig.height=5}
plot(
  pleasure_ftna,
  type = "combined",
  network = "summary",
  ribbon = TRUE,
  show_weights = TRUE,
  layout = "circle"
)
```

## `ts_tna()`: transition probabilities

`ts_tna()` row-normalizes the adjacent counts. Rows are source states, columns
are next states, and nonempty rows sum to one.

```{r ts-tna, eval=has_nestimate}
pleasure_tna <- ts_tna(pleasure_states)

class(pleasure_tna)
round(pleasure_tna$weights, 3)
rowSums(pleasure_tna$weights)
```

### Combined plot

```{r ts-tna-combined, eval=has_nestimate && has_cograph, fig.width=11, fig.height=5}
plot(
  pleasure_tna,
  type = "combined",
  network = "summary",
  ribbon = TRUE,
  show_weights = TRUE,
  layout = "circle"
)
```

## `ts_cna()`: co-occurrence network

`ts_cna()` counts state co-occurrence within the sequence. It discards
direction, so its matrix is symmetric. Its weights are counts, not
probabilities.

```{r ts-cna, eval=has_nestimate}
pleasure_cna <- ts_cna(pleasure_states)

class(pleasure_cna)
pleasure_cna$weights
isSymmetric(pleasure_cna$weights)
```

### Combined plot

```{r ts-cna-combined, eval=has_nestimate && has_cograph, fig.width=11, fig.height=5}
plot(
  pleasure_cna,
  type = "combined",
  network = "summary",
  ribbon = TRUE,
  show_weights = FALSE,
  layout = "circle"
)
```

## `ts_atna()`: attention transition network

`ts_atna()` applies Nestimate's attention weighting to directed relationships
between states. The raw weights are not probabilities.

```{r ts-atna, eval=has_nestimate}
pleasure_atna <- ts_atna(pleasure_states)

class(pleasure_atna)
round(pleasure_atna$weights, 2)
```

### Combined plot

```{r ts-atna-combined, eval=has_nestimate && has_cograph, fig.width=11, fig.height=5}
plot(
  pleasure_atna,
  type = "combined",
  network = "summary",
  ribbon = TRUE,
  show_weights = FALSE,
  layout = "circle"
)
```

# `series_networks()`

`series_networks()` splits a pooled transition model into one model for each
input series. This tutorial supplied one raw pleasure series, so the honest
result is a one-element list named `series_1`.

```{r series-networks, eval=has_nestimate}
pleasure_series_networks <- series_networks(pleasure_tna)

class(pleasure_series_networks)
length(pleasure_series_networks)
names(pleasure_series_networks)
pleasure_series_networks$series_1$weights
```

### Combined plot

This final combined plot is drawn from the model returned by
`series_networks()`.

```{r series-networks-combined, eval=has_nestimate && has_cograph, fig.width=11, fig.height=5}
plot(
  pleasure_series_networks$series_1,
  type = "combined",
  network = "summary",
  ribbon = TRUE,
  show_weights = TRUE,
  layout = "circle"
)
```

# Function checklist

| Export | Demonstrated result |
|---|---|
| `discretize()` | Low, Middle, and High pleasure states |
| `trend()` | Rolling pleasure direction |
| `vg()` | Horizontal visibility network |
| `tsn()` | Nearest-neighbor window-distance network |
| `ts_ftna()` | Raw transition counts plus combined plot |
| `ts_tna()` | Conditional transition probabilities plus combined plot |
| `ts_cna()` | Undirected co-occurrence counts plus combined plot |
| `ts_atna()` | Attention-weighted transitions plus combined plot |
| `series_networks()` | One model for the one supplied series plus combined plot |

The functions describe different properties of the same unchanged-order
pleasure segment. None of these representations creates independent
participants or supports an inferential claim by itself.
