---
title: "From Time Series to Networks"
author:
  - name: "Mohammed Saqr"
    affiliation: "University of Eastern Finland"
  - name: "Sonsoles López-Pernas"
    affiliation: "University of Eastern Finland"
  - name: "Manuel J. Gómez"
    affiliation: "University of Murcia"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{From Time Series to Networks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 150,
  out.width = "100%"
)
library(tsn)
has_cograph <- requireNamespace("cograph", quietly = TRUE) &&
  requireNamespace("igraph", quietly = TRUE)
has_nestimate <- requireNamespace("Nestimate", quietly = TRUE)
```

# Introduction

Time series analysis traditionally represents temporal data as observations
indexed by time, with attention directed to quantities such as level,
variation, dependence, and change. A network representation offers a
complementary view: it replaces, or augments, the explicit time index with a
relational structure in which nodes stand for observations, temporal states, or
local windows, and edges encode a defined relationship between them. That
relationship may be geometric visibility, numerical similarity, a state
transition, a co-occurrence, or a weighted dependence. A time series therefore
does not have a single canonical network. The network that results is
determined jointly by the chosen representation and by the structural question
that representation is designed to answer.

The **tsn** package provides a set of these representations while keeping the
modelling choices explicit at every step. `discretize()` maps continuous
observations onto a finite set of ordered states and so yields a symbolic
version of the series. `trend()` characterizes local direction and separates
short-range increase, decrease, stability, and turbulence. `vg()` constructs a
visibility graph, in which observations are joined when the geometry of the
series makes them mutually visible. `tsn()` is the general distance-network
constructor over complete series, local windows, or individual observations;
this vignette applies it to overlapping windows joined by their similarity.

Once a series has been reduced to a sequence of states, its dynamics can be
examined through transition network analysis. `ts_tna()` builds a transition
network whose edges carry conditional transition probabilities. `ts_ftna()`
builds the corresponding frequency network, `ts_cna()` a co-occurrence network,
and `ts_atna()` an attention-weighted transition network. These four
constructions summarize different properties of the same symbolic sequence and
are complementary rather than interchangeable. Finally, `series_networks()`
extracts and indexes the network model attached to each source series,
retaining the model type, the shared state alphabet, and the provenance needed
to trace a network back to its data.

Every representation in this vignette is developed on one fixed segment of one
measured variable, so that the constructions can be compared without any change
in the underlying data. Network figures are produced by the package's `plot()`
methods through **cograph**, and the four transition-network constructors
require the **Nestimate** package; both are optional and used only where noted.

| Scientific question | Function | Result | What it captures |
|---|---|---|---|
| **How can a continuous series be represented as a sequence of discrete states?** | `discretize()` | Tidy state table | Maps continuous observations onto a finite, ordered state alphabet, making recurring levels and ranges explicit. |
| **How does the series change locally over time?** | `trend()` | Tidy trend table | Labels each observation by its local direction: increase, decrease, stability, or turbulence. |
| **Which observations are connected by geometric visibility?** | `vg()` | Visibility network | Connects observations that can "see" each other over the intervening values. |
| **Which time windows are similar in value or shape?** | `tsn()` | Distance network | Relates local windows by their proximity, exposing repeated or structurally similar segments. |
| **How likely is one state to be followed by another?** | `ts_tna()` | Transition network (TNA) | Weights directed edges by the conditional probability of each successive transition. |
| **How often does the series move from one state to another?** | `ts_ftna()` | Frequency network (FTNA) | Weights directed edges by the observed count of each successive transition. |
| **Which states co-occur within the sequence?** | `ts_cna()` | Co-occurrence network (CNA) | Uses undirected edges to represent state co-occurrence, ignoring direction. |
| **Which transitions receive the greatest attention weight?** | `ts_atna()` | Attention-weighted network (ATNA) | Weights directed edges by model-assigned attention rather than raw counts. |
| **How is each network associated with its source series?** | `series_networks()` | Indexed model collection | Organizes extracted models by source series while preserving their correspondence and provenance. |

In short, `discretize()` defines the **states**, `trend()` describes **local
direction**, `vg()` describes **geometric visibility**, and `tsn()` describes
**similarity between temporal windows**. The `ts_tna()` family then builds
transition networks from the state sequence, differing only in the quantity
assigned to their edges, and `series_networks()` indexes the results by source
series.

# Data

The `motivation` data set bundled with the package is an example of
experience-sampling, or intensive longitudinal, data: the same respondent was
prompted several times a day and asked to report, in the moment, how they felt
about the activity they were engaged in. Each of its 4,871 rows is one such
momentary report, collected between September 2018 and July 2021, and no values
are missing. The thirteen columns fall into a few natural groups. Three record
motivation (`autonomy`, `competence`, `relatedness`); three record appraisal of
the current activity (`pleasure`, `interest`, `importance`); several describe
the situation and its emotional framing, together with a categorical context
label whose values are Home, Work, Personal, and Other; and one records overall
`mood`. Two further columns place each report in time, giving the calendar date
and the position of the report within that day's sequence of prompts.

This vignette follows a single one of these signals, the `pleasure` rating,
which captures how pleasant the respondent found the current activity at the
moment of the prompt. To keep every network small enough to read in full, the
analysis uses only the first 60 reports, which each verb selects by name from
the data frame. Passing the data frame rather than a bare vector preserves the
recorded order of the observations, and that order is what every method below
treats as the flow of time.

```{r data}
data("motivation", package = "tsn")

pleasure <- head(motivation, 60)

dim(pleasure)
```

These 60 reports were collected over 21 days, from 29 September to 19 October
2018. The pleasure ratings occupy the observed range 6 to 44, with a median of
30, a mean of 26.8, and a standard deviation of about 10, so they spread across
a wide band rather than clustering near a single value. The segment carries no
systematic trend: the first twenty reports average 27.5 and the last twenty
average 27.0, essentially unchanged. The representations that follow should
therefore be read as descriptions of short-run fluctuation around a roughly
constant level, not of a rising or falling series. The distinction matters for
interpretation, because several of the constructions below, in particular the
visibility graph and the trend labels, respond to the local rises and falls of
the series rather than to its overall direction.

# Transition Network Analysis

Transition Network Analysis starts by turning the observed series into a finite
sequence of states. The aim is not to discard the measured values but to obtain
a symbolic representation in which dynamics can be studied as movement among a
fixed set of states. `discretize()` performs this step, assigning each
observation to one state and returning one tidy row per observation.

Here the default quantile discretization splits the values into three empirical
groups, labelled `Low`, `Middle`, and `High`. The resulting sequence keeps the
temporal order of the observations and replaces their continuous values with
ordered categorical states.

```{r discretize}
states <- discretize(
  data = pleasure,
  series = "pleasure",
  labels = c("Low", "Middle", "High")
)

states
summary(states)
```

Each row records the source identifier, the time index, the observed value, the
assigned state, and an assignment probability. The `probability` column reports
the confidence of the assignment: it equals 1 for a hard partition such as the
quantile rule used here, and takes values in the unit interval for probabilistic
discretizers such as `gaussian` or `kde`. Because the state labels follow the
ordering of the values, `Low`, `Middle`, and `High` retain a strict ordinal
meaning. Quantile boundaries do not, in general, divide the observations into
exactly equal groups when values are tied; in this segment the counts are 20
`Low`, 23 `Middle`, and 17 `High`.

The state sequence is the intermediate object from which every transition
network below is built. Its three states become the network nodes, and each
adjacent pair of observations defines one possible state-to-state transition.
The four constructors that follow all read this same sequence; what differs
between them is the quantity placed on the edges.

```{r discretize-plot, fig.alt="Ribbon plot of the pleasure series with a strip of Low, Middle, and High states drawn beneath the measured values along observation order."}
plot(states, type = "ribbon")
```

The ribbon view places the discrete states beneath the measured series, so the
correspondence between an observation and its assigned state stays visible
without a full-panel overlay obscuring the values themselves.

## Transition probabilities with `ts_tna()`

`ts_tna()` builds a directed network in which an edge runs from the current
state to the next, weighted by the conditional probability of that transition.
Each row of the weight matrix is a probability distribution over the possible
next states, so the outgoing weights of a state sum to one.

```{r ts-tna, eval=has_nestimate}
probabilities <- ts_tna(states)

probabilities
```

The diagonal entries measure persistence, the probability that the next
observation stays in the current state: 0.300 for `Low`, 0.409 for `Middle`,
and 0.294 for `High`. Comparing these with the marginal state proportions
(0.333, 0.383, and 0.283) separates persistence from mere prevalence. `Middle`
and `High` both persist above their marginal share, so a `Middle` or `High`
observation is more likely to be followed by its own state than random ordering
would imply. `Low` sits marginally below its share (0.300 against 0.333),
indicating no such tendency to persist over this short segment.

```{r ts-tna-plot, eval=has_nestimate && has_cograph, fig.width=9, fig.alt="Directed transition-probability network over the Low, Middle, and High states, with arrows labelled by conditional probabilities."}
plot(probabilities)
```

Row normalization is what makes this representation interpretable. Each row is
the conditional distribution of the next state given the current one, so the
outgoing patterns of different states remain comparable even though the states
have different marginal frequencies. A sequence of 60 observations is thereby
reduced to three nodes and the directed relationships among them, with the
temporal order retained as the basis for the estimates.

## Transition frequencies with `ts_ftna()`

`ts_ftna()` weights each directed edge by the observed count of the
corresponding transition, without the row normalization applied by `ts_tna()`.
The network therefore preserves the absolute number of times each transition
occurred.

```{r ts-ftna, eval=has_nestimate}
counts <- ts_ftna(states)

counts
```

The 60 observations form 59 adjacent pairs, so the count matrix sums to 59. The
most frequent single transition is `Middle` to `Middle`, which occurs 9 times
and is the only self-transition that is also the largest count in its row. The
`Low` and `High` rows are more evenly spread: a `Low` observation is followed
equally often by `Middle` or `High` (7 each) rather than by another `Low` (6),
so persistence, although present, does not dominate the raw counts here. The
frequency and probability networks describe the same transitions but answer
different questions: the frequency network reports **how often** each transition
occurred, whereas the probability network reports **how likely** it is given the
current state.

```{r ts-ftna-plot, eval=has_nestimate && has_cograph, fig.width=9, fig.alt="Directed frequency transition network over the three states, with arrows labelled by observed transition counts."}
plot(counts)
```

## State co-occurrence with `ts_cna()`

`ts_cna()` builds an undirected, symmetric network whose edges measure how
strongly two states occur together, without regard to the order in which they
appear. It answers a different question from the directed transition networks.

```{r ts-cna, eval=has_nestimate}
cooccurrence <- ts_cna(states)

cooccurrence
```

For a single sequence, these co-occurrence counts follow directly from the
marginal state frequencies. An off-diagonal entry is the product of the two
state counts, so the `Low`-`Middle` weight is `20 * 23 = 460`; a diagonal entry
counts the unordered pairs within a state, so the `Low`-`Low` weight is
`20 * 19 / 2 = 190`. The network therefore reflects the composition of the state
alphabet rather than the temporal direction of transitions. A strong edge means
the two states are both common, not that one tends to follow the other.

```{r ts-cna-plot, eval=has_nestimate && has_cograph, fig.width=9, fig.alt="Undirected co-occurrence network over the three states, with symmetric edges weighted by state co-occurrence."}
plot(cooccurrence)
```

The symmetric drawing makes the distinction explicit: an edge carries no
source-to-target direction, in contrast to the transition networks above.

## Attention-weighted transitions with `ts_atna()`

`ts_atna()` assigns each transition an attention weight rather than a count or a
probability. The edge weights are accumulated attention values and are not on
the scale of either the frequency or the probability network.

```{r ts-atna, eval=has_nestimate}
attention <- ts_atna(states)

attention
```

The weights range from 2.875 to 5.015. Unlike the frequency and probability
networks, the largest weights here are off-diagonal: the strongest
attention-weighted transition is `Middle` to `Low` (5.015), and the persistence
diagonals are not the dominant entries in their rows. The attention mechanism
consequently redistributes emphasis away from the self-transitions that the raw
counts favour, and the weights should be read on the model's own attention
scale rather than compared directly with counts or probabilities.

```{r ts-atna-plot, eval=has_nestimate && has_cograph, fig.width=9, fig.alt="Directed attention-weighted transition network over the three states, with edges weighted by accumulated attention values."}
plot(attention)
```

## Indexing the model with `series_networks()`

`series_networks()` returns an indexed collection of the network models attached
to the source series. It is designed for analyses that extract networks across
many series at once, where the index preserves the link between each series and
its model. With a single input series it returns one entry, corresponding to
`pleasure`.

```{r series-networks, eval=has_nestimate}
individual <- series_networks(probabilities)

individual
```

The index records the source series, the network type, and the number of
observations, states, and edges. For the transition-probability network the
three states generate nine directed relationships, the six between distinct
states plus the three self-transitions.

```{r series-networks-plot, eval=has_nestimate && has_cograph, fig.width=7, fig.alt="The single indexed transition-probability network for the pleasure series, drawn from the series_networks collection."}
plot(individual)
```

The collection thus provides a higher-level handle on the extracted networks
while keeping their provenance. Given several source series, the same structure
organizes and compares their models systematically.

# Local direction with `trend()`

`trend()` estimates a rolling local slope and labels each observation as
`Ascending`, `Descending`, `Flat`, `Turbulent`, `Missing Data`, or `Initial`.
The window is adaptive and defaults to one tenth of the series length, which is
6 observations here, and the default Theil-Sen slope limits the influence of
isolated extreme ratings.

```{r trend}
directions <- trend(
  data = pleasure,
  series = "pleasure"
)

directions
summary(directions)
```

`Ascending` is the most common label with 21 observations, followed by
`Descending` with 17, `Turbulent` with 13, `Initial` with 5, and `Flat` with 4.
The near balance between ascending and descending positions is what a segment
without drift should produce: local direction keeps reversing rather than
accumulating in one direction. `Initial` marks the leading positions that lack a
complete centered window, and `Missing Data` would mark absent values, of which
this segment has none.

```{r trend-plot, fig.alt="The pleasure series with each observation coloured by its trend label: Ascending, Descending, Flat, Turbulent, or Initial."}
plot(directions)
```

The default plot colours each observation by its trend label.

```{r trend-panels, fig.alt="Two stacked panels: the pleasure series on top and its rolling slope metric below, with a shaded flat band marking near-zero slope."}
plot(directions, type = "panels")
```

The panel view places the rolling metric below the measurements, so the points
at which the metric crosses the flat band can be checked against the assigned
direction.

# Visibility with `vg()`

`vg()` treats each observation as a node. Under natural visibility, two points
are joined when the straight line between their values stays above every
intervening point. Under horizontal visibility, they are joined only when every
intervening value lies below both endpoints, a strictly stronger condition.

```{r vg-natural}
natural <- vg(
  data = pleasure,
  series = "pleasure"
)

summary(natural)
```

The natural visibility graph has 60 nodes and 143 edges.

```{r vg-horizontal}
horizontal <- vg(
  data = pleasure,
  type = "horizontal",
  series = "pleasure"
)

summary(horizontal)
```

The stricter horizontal rule retains 107 of those connections. In both graphs
the edge weights are binary, since visibility either holds or it does not.

At this length individual edges remain distinguishable, so the horizontal graph
can be drawn directly.

```{r vg-network, eval=has_cograph, fig.alt="Horizontal visibility graph of sixty pleasure observations, drawn as a network with nodes sized by degree and the temporal chain visible alongside longer visibility edges."}
plot(horizontal)
```

The temporal chain is visible alongside the longer edges that reach across lower
intervening observations. Node size encodes degree; peaks accumulate more
connections because they stay visible over the smaller values around them.

```{r vg-series, fig.alt="The sixty pleasure observations behind the visibility graph, plotted as a line over observation order."}
plot(horizontal, type = "series")
```

The source-series view recovers the measurements behind the network object and
makes the short-run fluctuation that the visibility rule reads directly
apparent.

# Window similarity with `tsn()`

`tsn()` is the general distance-network constructor. With `method = "distance"`
a single measured series defaults to sliding-window nodes. The window width
defaults to one tenth of the series length, which is 6 here, and a step of 2
yields 28 overlapping windows.

```{r tsn-distance}
windows <- tsn(
  data = pleasure,
  method = "distance",
  series = "pleasure",
  step = 2
)

summary(windows)
```

The default full connection rule evaluates and keeps all 378 window pairs. Each
edge weight is `1 / (1 + distance)`, so larger weights mark more similar
windows.

```{r tsn-nearest}
sparse <- tsn(
  data = pleasure,
  method = "distance",
  series = "pleasure",
  step = 2,
  connect = "nearest",
  neighbors = 2
)

summary(sparse)
```

Nearest-neighbour sparsification keeps 40 edges and lowers the density from 1 to
0.106, retaining the strongest local similarities instead of drawing every pair.

```{r tsn-network, eval=has_cograph, fig.alt="Nearest-neighbour distance network over twenty-eight sliding windows of the pleasure series, with edges joining the most similar windows."}
plot(sparse)
```

No retained edge joins immediately adjacent windows. The closest retained pairs
are two windows apart and the most distant are twenty-one windows apart, which
shows that overlap in time does not by itself determine similarity in value:
windows far apart in the sequence can still be each other's nearest neighbours.
