---
title: "Time Series Networks"
author: "tsn package authors"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Time Series Networks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 150,
  out.width = "100%"
)
library(tsn)
has_cograph <- requireNamespace("cograph", quietly = TRUE) &&
  requireNamespace("igraph", quietly = TRUE)
has_nestimate <- requireNamespace("Nestimate", quietly = TRUE)
```

# Introduction

Time series analysis traditionally represents temporal data through observations
indexed by time, with emphasis placed on quantities such as level, variation,
dependence, and temporal change. A network-based representation provides a
complementary perspective by replacing, or augmenting, the explicit time index
with a relational structure. In this framework, observations, temporal states,
or local windows can be represented as nodes, while edges encode relationships
defined by geometric visibility, similarity, state transitions, co-occurrence,
or weighted dependence. The resulting network is therefore not a unique
property of a time series. It is determined by the representation used and by
the structural question that the representation is designed to answer.

The **tsn** R package implements a collection of network representations for
time series while keeping these modelling choices explicit. The `discretize()`
function maps continuous observations to a finite set of states, providing a
symbolic representation of the series. The `trend()` function characterizes
local direction and captures short-range increases, decreases, and stable or
turbulent behaviour. The `vg()` function constructs a visibility graph in which
observations are connected according to their geometric visibility. The `tsn()`
function constructs distance networks over complete series, local windows, or
individual observations. This vignette uses local windows and connects them
according to their similarity.

Once a time series has been represented as a sequence of discrete states, its
temporal dynamics can be studied through Transition Network Analysis (TNA). The
`ts_tna()` function constructs a Transition Network (TNA). The `ts_ftna()`
function constructs a Frequency Transition Network (FTNA), `ts_cna()` constructs
a Co-occurrence Network (CNA), and `ts_atna()` constructs an Attention-Weighted
Transition Network (ATNA). These constructions emphasize different properties
of the same symbolic sequence and should be interpreted as complementary rather
than interchangeable models.

The package also provides `series_networks()` for extracting and organizing the
network model associated with each source series. The function retains the
model type, shared state alphabet, provenance, and source index. In this
single-series vignette, it returns one indexed model and provides a direct way to
print, summarize, or plot that model.

This vignette develops every representation using one unchanged segment of one
measured variable. The examples show how each function is called and how the
resulting structure can be interpreted. Network figures are generated through
the package's `plot()` methods using cograph. The four transition-network
constructors require the Nestimate package.

The functions and their returned representations are:

| Scientific question | Function | Result | What it captures |
|---|---|---|---|
| **How can the continuous series be represented as a sequence of discrete states?** | `discretize()` | Tidy state table | Converts continuous observations into a finite state representation, making recurring levels, ranges, or regimes explicit. |
| **How does the series change locally over time?** | `trend()` | Tidy trend table | Captures local directional behaviour, distinguishing periods of increase, decrease, stability, or turbulence. |
| **Which observations are connected by geometric visibility?** | `vg()` | Visibility network | Captures the geometric organization of the series by connecting observations according to their visibility relationships. |
| **Which time windows show similar values or patterns?** | `tsn()` | Distance network | Represents relationships among local windows according to their proximity or similarity, revealing repeated or structurally similar temporal patterns. |
| **How likely is one state to transition to another?** | `ts_tna()` | Transition Network (TNA) | Weights directed edges between states by the conditional probability of each successive transition. |
| **How frequently does the series transition from one state to another?** | `ts_ftna()` | Frequency Transition Network (FTNA) | Weights directed edges between states by the observed frequency of each successive transition. |
| **Which states are connected through co-occurrence in the sequence?** | `ts_cna()` | Co-occurrence Network (CNA) | Uses undirected edges to represent state co-occurrence within the sequence. |
| **Which transitions receive greater attention weight?** | `ts_atna()` | Attention-Weighted Transition Network (ATNA) | Weights directed edges between states by their attention values. |
| **How are network representations associated with their source series?** | `series_networks()` | Indexed model collection | Organizes extracted network models by source series while preserving their correspondence and provenance. |

The conceptual distinction is therefore clear. `discretize()` defines the
**states**, `trend()` describes **local direction**, `vg()` describes
**geometric visibility**, and `tsn()` describes **similarity between temporal
windows**. The `ts_tna()` family constructs transition-network representations
from the resulting sequence, differing in the information assigned to their edges.
`series_networks()` organizes the extracted models by source series.

# Data

The `motivation` data set contains 4,871 intensive longitudinal observations.
The analysis uses the first 60 rows and selects the `pleasure` measurement
inside each tsn verb. Keeping the data as a data frame avoids manual column
extraction and preserves its original row order.

```{r data}
data("motivation", package = "tsn")

pleasure <- head(motivation, 60)

dim(pleasure)
```

The selected ratings range from 6 to 44. Their mean is 26.8 and their median
is 30. The first twenty observations average 27.5 and the final twenty average
27.0, so this short segment carries no pronounced drift. The later plots should
therefore be read as short-run fluctuation around a roughly stable level rather
than as a trending series.

# Transition Network Analysis

Transition network analysis begins by transforming the observed time series
into a finite sequence of states. The purpose of this transformation is not to
replace the measured values, but to provide a symbolic representation in which
temporal dynamics can be examined as transitions among a defined set of states.
`discretize()` performs this transformation by assigning each observation to a
state and returning one tidy row per observation.

For this example, the default quantile-based discretization divides the observed
values into three empirical groups. The state labels are supplied as `Low`,
`Middle`, and `High`. Thus, the resulting sequence preserves the temporal
ordering of the observations while replacing their continuous values with
categorical states.

```{r discretize}
states <- discretize(
  data = pleasure,
  series = "pleasure",
  labels = c("Low", "Middle", "High")
)

states
summary(states)
```

The resulting object retains the source identifier, time index, observed value,
assigned state, and assignment probability. The state labels are ordered
according to the corresponding values, so `Low`, `Middle`, and `High` retain
their ordinal interpretation. Because the discretization is based on empirical
quantiles, ties at the boundaries can prevent the three categories from
containing exactly the same number of observations. Here, the resulting state
frequencies are 20 `Low`, 23 `Middle`, and 17 `High`.

The state sequence is the intermediate representation from which the transition
networks are constructed. The three states define the network nodes, while
successive observations in the sequence define the possible state-to-state
transitions. The different transition-network constructors applied below use
this same state sequence; what changes is the quantity used to represent the
relationships between states.

```{r discretize-plot}
plot(states, type = "ribbon")
```

The ribbon representation displays the state sequence beneath the measured
series. This preserves the temporal correspondence between the original
observations and their discrete representation without obscuring the measured
values with a full-panel categorical overlay.

## Transition Network with `ts_tna()`

`ts_tna()` constructs a transition network from the discrete state sequence in
which directed edges represent transitions from the current state to the
subsequent state. Edge weights are conditional transition probabilities. For a
given source state, the outgoing edge weights therefore describe the
distribution of possible next states.

```{r ts-tna, eval=has_nestimate}
probabilities <- ts_tna(states)

probabilities
```

The diagonal entries quantify persistence, that is, the probability that the
next observation remains in the current state. In this example, the
self-transition probabilities are 0.300 for `Low`, 0.409 for `Middle`, and
0.294 for `High`. Each exceeds the corresponding marginal state proportion,
indicating that consecutive observations are more likely to remain in the same
state than the marginal frequencies alone would imply.

```{r ts-tna-plot, eval=has_nestimate && has_cograph, fig.width=9}
plot(probabilities)
```

Row normalization is central to the interpretation of this representation. Each
row describes the conditional distribution of the next state given the current
state, making outgoing transition patterns directly comparable even when the
states have different marginal frequencies.

The combined representation links the original state sequence to the resulting
transition network. A sequence of 60 observations is therefore reduced to
three state nodes and the directed relationships connecting them, while the
temporal ordering remains the basis for estimating those relationships.

## Frequency Transition Network (FTNA) with `ts_ftna()`

`ts_ftna()` constructs a transition network using the observed frequency of
each state-to-state transition. Unlike `ts_tna()`, the edge weights are not
normalized by the number of transitions originating from each state. The
resulting network therefore retains information about the absolute frequency
with which transitions occur.

```{r ts-ftna, eval=has_nestimate}
counts <- ts_ftna(states)

counts
```

With 60 observations, there are 59 adjacent observation pairs and consequently
59 observed transitions. The transition-count matrix therefore sums to 59.
The diagonal entries are the largest entries in each row, indicating that
remaining in the current state occurs more frequently than transitioning to
either of the alternative states.

The frequency network and the probability network describe the same underlying
transitions but answer different questions. The frequency network measures
**how often** each transition occurred, whereas the probability network measures
**how likely** each transition is conditional on the current state.

```{r ts-ftna-plot, eval=has_nestimate && has_cograph, fig.width=9}
plot(counts)
```

## Co-occurrence Transition Network with `ts_cna()`

`ts_cna()` constructs a transition network based on the co-occurrence of states
within the transition sequence. Unlike the directed transition representation
produced by `ts_tna()` and `ts_ftna()`, the resulting network is undirected and
symmetric. It therefore describes the strength of association between states
without distinguishing the direction in which the states occur.

```{r ts-cna, eval=has_nestimate}
cooccurrence <- ts_cna(states)

cooccurrence
```

For a single state sequence, these co-occurrence counts are determined by the
marginal state frequencies. For example, the `Low`–`Middle` relationship is
`20 × 23 = 460`, whereas the `Low`–`Low` diagonal is
`20 × 19 / 2 = 190`. The resulting network consequently captures the
association implied by state composition rather than the temporal direction of
successive transitions.

This distinction is important when interpreting the network. A strong edge in
the co-occurrence representation does not indicate that one state tends to
follow another. It indicates that the two states occur together within the
defined co-occurrence structure.

```{r ts-cna-plot, eval=has_nestimate && has_cograph, fig.width=9}
plot(cooccurrence)
```

The symmetric network representation makes this distinction visually explicit:
unlike the directed transition networks, there is no source-to-target direction
associated with an edge.

## Attention-Weighted Transition Network (ATNA) with `ts_atna()`

`ts_atna()` constructs a transition network in which transitions are assigned
attention weights rather than being represented solely by their raw frequencies
or normalized probabilities. The resulting edge weights are accumulated
attention values and therefore should not be interpreted as either transition
counts or probabilities.

```{r ts-atna, eval=has_nestimate}
attention <- ts_atna(states)

attention
```

The resulting weights range from 2.875 to 5.015. As in the frequency and
probability networks, the diagonal entries are the largest outgoing
relationships for all three states, indicating that persistence remains
prominent under the attention-weighted representation.

The important distinction is that the edge magnitude now reflects the attention
assigned to a transition by the model rather than simply the number of times the
transition was observed. Consequently, attention weights should be interpreted
according to the weighting mechanism of the `ts_atna()` model rather than on the
scale of the frequency or probability networks.

```{r ts-atna-plot, eval=has_nestimate && has_cograph, fig.width=9}
plot(attention)
```

## Extracting the Network Model with `series_networks()`

`series_networks()` provides an indexed collection of the network models
associated with the source series. It is intended for situations in which
network representations are extracted systematically across multiple series.
The index preserves the relationship between each source series and its
corresponding network model.

In this example, only one source series is supplied, so the resulting collection
contains a single entry corresponding to `pleasure`.

```{r series-networks, eval=has_nestimate}
individual <- series_networks(probabilities)

individual
```

The resulting index records the source series, network type, number of
observations, number of states, and number of edges. For the
transition-probability network, the three states generate nine possible directed
state-to-state relationships, including the three self-transitions.

```{r series-networks-plot, eval=has_nestimate && has_cograph, fig.width=7}
plot(individual)
```

The resulting model collection therefore provides a higher-level representation
of the extracted networks while retaining their provenance. With multiple
source series, the same structure can be used to organize and compare the
corresponding network models systematically.

# Local direction with `trend()`

`trend()` estimates a rolling local direction and labels each observation as
Ascending, Descending, Flat, Turbulent, Missing Data, or Initial. Its adaptive
window is 6 observations for this segment. The default Theil-Sen slope limits
the influence of isolated extreme ratings.

```{r trend}
directions <- trend(
  data = pleasure,
  series = "pleasure"
)

directions
summary(directions)
```

Ascending is the most frequent class, with 21 observations. Descending covers
17 observations, Turbulent 13, Initial 5, and Flat 4. Ascending and descending
positions are close in number, which is what a segment without pronounced drift
should produce: local direction keeps reversing rather than accumulating.
Initial marks positions without a complete centered rolling window.

```{r trend-plot}
plot(directions)
```

The default plot colours each observation by its trend class.

```{r trend-panels}
plot(directions, type = "panels")
```

The panel view places the rolling metric below the measurements. Crossings of
the flat band can therefore be checked against the assigned direction.

# Visibility with `vg()`

`vg()` makes each observation a node. Natural visibility connects two points
when the line between their values passes above every intervening point.
Horizontal visibility connects them when every intervening value is below both
endpoints.

```{r vg-natural}
natural <- vg(
  data = pleasure,
  series = "pleasure"
)

summary(natural)
```

The natural visibility graph contains 60 nodes and 143 edges.

```{r vg-horizontal}
horizontal <- vg(
  data = pleasure,
  type = "horizontal",
  series = "pleasure"
)

summary(horizontal)
```

The stricter horizontal rule retains 107 edges. Both constructions have binary
edge weights because visibility either holds or does not hold.

At this length the graph is small enough for individual edges to stay
distinguishable, so the horizontal construction can be drawn directly.

```{r vg-network, eval=has_cograph}
plot(horizontal)
```

The temporal chain remains visible together with longer edges that cross lower
intervening observations. Node size represents degree. Peaks tend to have more
connections because they remain visible across smaller values.

```{r vg-series}
plot(horizontal, type = "series")
```

The source-series view retains the measurements behind the network object. It
also makes the short-run fluctuation that the visibility rule reads visible.

# Window similarity with `tsn()`

`tsn()` is the general constructor. With `method = "distance"`, one measured
series defaults to sliding-window nodes. The window width defaults to one tenth
of the series length and is therefore 6 here. A step of 2 produces 28
overlapping windows.

```{r tsn-distance}
windows <- tsn(
  data = pleasure,
  method = "distance",
  series = "pleasure",
  step = 2
)

summary(windows)
```

The default full connection rule evaluates and retains all 378 window pairs.
Edge weight is `1 / (1 + distance)`, so larger weights identify more similar
windows.

```{r tsn-nearest}
sparse <- tsn(
  data = pleasure,
  method = "distance",
  series = "pleasure",
  step = 2,
  connect = "nearest",
  neighbors = 2
)

summary(sparse)
```

Nearest-neighbour sparsification retains 40 edges and reduces density from 1 to
0.106. This version exposes the strongest local similarity structure instead of
drawing every possible pair.

```{r tsn-network, eval=has_cograph}
plot(sparse)
```

No retained edge joins immediately adjacent windows. The closest retained pairs
are two positions apart and the most distant are separated by 21 positions,
showing that overlap in time does not by itself determine Euclidean similarity.
