---
title: "Plotting Time Series Networks"
author:
  - name: "Mohammed Saqr"
    affiliation: "University of Eastern Finland"
  - name: "Sonsoles López-Pernas"
    affiliation: "University of Eastern Finland"
  - name: "Manuel J. Gómez"
    affiliation: "University of Murcia"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Plotting Time Series Networks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 110,
  out.width = "100%"
)
library(tsn)
has_cograph <- requireNamespace("cograph", quietly = TRUE) &&
  requireNamespace("igraph", quietly = TRUE)
has_nestimate <- requireNamespace("Nestimate", quietly = TRUE)
```

# Introduction

Which plot is drawn from a time series network selects which aspect of the data
becomes visible: the measured temporal sequence, the states assigned to
observations, or the relational structure defined by the network. These views
describe different aspects of the same analysis and should not be interpreted as
equivalent representations.

The plotting methods in **tsn** select the representation with the `type`
argument. Additional arguments control the visual encoding within that
representation, including state placement, point display, colour, labels, node
size, and edge annotation. When `type` is omitted, `plot()` uses the default type
for the corresponding result class.

This vignette uses one momentary `anxiety` series throughout. Each analytical object is
constructed immediately before its first plot, keeping the relationship between
the analysis and its graphical representation explicit.

| Result class | Plot types | Default |
|---|---|---|
| Discrete states from `discretize()` | `"overlay"`, `"ribbon"`, `"heatmap"`, `"stack"` | `"overlay"` |
| Local direction from `trend()` | `"points"`, `"ribbon"`, `"heatmap"`, `"panels"` | `"points"` |
| Network from `tsn()` or `vg()` | `"network"`, `"series"` | `"network"` |
| TNA-family model | `"combined"`, `"network"`, `"series"` | `"combined"` |
| Model collection from `series_networks()` | `"network"` | `"network"` |

For network objects, `"network"` displays the nodes and edges defined by the
model, whereas `"series"` shows the source time series from which the network
was constructed, with no network structure drawn over it. For TNA-family models,
`"combined"` displays the state
sequence together with the corresponding transition network, while `"network"`
focuses on the state-to-state relationships alone. The choice of plot type
therefore determines whether the figure is used to examine the temporal
sequence, the state representation, or the network structure.

# Data

The examples use the first 60 momentary anxiety reports of one student,
Jamal, from the bundled `esm_srl` experience-sampling data (41 students,
0-100 scales, fully anonymized). A 25-observation segment is used for
visibility analysis so that individual nodes and edges remain
distinguishable.

```{r data}
data("esm_srl", package = "tsn")

anxiety <- head(subset(esm_srl, name == "Jamal"), 60)
short <- head(anxiety, 25)
```

# Discrete-state plots

The `plot()` method for a `tsn_states` object provides four representations of
the same discretization.

| `type` | Graphical encoding | Scientific use |
|---|---|---|
| `"overlay"` | State value bands behind the measured series | Relating assigned states to their value ranges and boundaries |
| `"ribbon"` | State strip beneath an unobscured series | Retaining exact variation when state changes are frequent |
| `"heatmap"` | One categorical tile per observation | Displaying the state sequence as a compact temporal signature |
| `"stack"` | Companion series aligned above the discretized series | Relating the states of a derived quantity to its source measurement |

The principal arguments are `type` and `series`. For overlays, `overlay`
selects horizontal ranges (the default), vertical regions, or no shading; `lines` can add
state boundaries; and `min_run` suppresses very short displayed runs without
changing the fitted states. The arguments `points`, `palette`, `alpha`,
`legend`, and `grid` control graphical appearance. Heatmaps additionally accept
`sort` and `border`. The stack view requires a numeric companion supplied with
`with`; `shade` and `color_line` determine how the state classification is
carried across the two panels.

## Overlay

The series is first discretized into three empirical quantile states and then
plotted. Because `"overlay"` is the default type, no plotting argument is
required.

```{r states-overlay}
states <- discretize(
  data = anxiety,
  series = "anxiety",
  labels = c("Low", "Middle", "High")
)

plot(states)
```

The horizontal bands identify the value ranges associated with the three states,
while the line retains the measured observations. This view emphasizes the
relationship between state membership, magnitude, and the fitted boundaries.

## Ribbon

```{r states-ribbon}
plot(states, "ribbon")
```

The ribbon places the state encoding outside the measurement panel. It is
preferable when vertical shading would conceal short-range variation.

## Heatmap

```{r states-heatmap, fig.height=2.5}
plot(states, "heatmap")
```

Each tile represents one time point and its colour represents the assigned
state. For one series, the result is a compact categorical timeline.

## Stack

The stack view is appropriate when states are defined from a transformation of
the measured series. Here, one-step changes are discretized and plotted beneath
the original observations.

```{r states-stack, fig.height=7}
changes <- with(anxiety, c(0, diff(anxiety)))

change_states <- discretize(
  data = changes,
  labels = c("Negative", "Stable", "Positive")
)

plot(
  change_states,
  "stack",
  with = with(anxiety, anxiety),
  with_label = "Anxiety"
)
```

The shared time axis permits changes in the original measurement to be read
directly against their categorical representation.

# Local-direction plots

The `plot()` method for a `tsn_trend` object provides four views of the rolling
direction classification.

| `type` | Graphical encoding | Scientific use |
|---|---|---|
| `"points"` | Observations coloured by direction | Locating ascending, descending, flat, and turbulent intervals |
| `"ribbon"` | Direction strip beneath the measured series | Separating classification from measurement variation |
| `"heatmap"` | Categorical direction tiles | Summarizing the timing of local direction |
| `"panels"` | Measured series above the rolling trend metric | Diagnosing how the numerical metric produces the classification |

The common arguments `series`, `palette`, `legend`, and `grid` control
selection and presentation. The line and point encodings are modified with
`line_color`, `line_width`, and `point_size`. In the panel view, `flat_band`
controls whether the classification threshold is shown. Heatmap presentation
can be modified with `sort` and `border`.

## Points

The local-direction model is fitted and immediately plotted using its default
point representation.

```{r trend-points}
directions <- trend(
  data = anxiety,
  series = "anxiety"
)

plot(directions)
```

Colour encodes the local trend state at each observation: *ascending*,
*descending*, and *flat* record the direction of the window centred on that
point, while *turbulent* marks intervals too volatile to read as a single
direction. The view retains individual measurements and is therefore suited to
locating directional changes precisely in time.

## Ribbon

```{r trend-ribbon}
plot(directions, "ribbon")
```

The ribbon is useful when adjacent classifications change frequently, because
the measured line remains visually continuous.

## Heatmap

```{r trend-heatmap, fig.height=2.5}
plot(directions, "heatmap")
```

The heatmap removes magnitude and preserves only temporal classification. It
should therefore be interpreted as a map of direction, not as a plot of the
observed values.

## Panels

```{r trend-panels, fig.height=6}
plot(directions, "panels")
```

The lower panel displays the rolling metric used to assign direction. This is
the appropriate diagnostic view for evaluating crossings of the flat band and
their correspondence with the measured series.

# Constructed-network plots

Objects returned by `tsn()` and `vg()` provide two plot types.

| `type` | Graphical encoding | Scientific use |
|---|---|---|
| `"network"` | Nodes and edges of the constructed network | Examining relational structure, central observations, and connectivity |
| `"series"` | Measurements retained in the network object | Verifying the temporal data from which the network was constructed |

For `type = "network"`, graphical arguments are passed to
`cograph::splot()`. Important options include `layout`, `labels`, `node_size`,
`node_fill`, `edge_color`, and edge-label controls. The tsn defaults use a
spring layout, degree-based node sizing, and unobtrusive edge styling. For
`type = "series"`, arguments such as `overlay`, `points`, `trend`, `legend`,
and `grid` modify the temporal view.

## Network

A horizontal visibility graph is constructed and immediately plotted. The
default type is `"network"`.

```{r network-view, eval=has_cograph}
visibility <- vg(
  data = short,
  type = "horizontal",
  series = "anxiety"
)

plot(visibility)
```

Nodes represent observations and edges represent horizontal visibility. The
network view removes the explicit time axis and emphasizes which observations
remain mutually visible across intervening values.

## Source series

```{r network-series}
visibility <- vg(
  data = short,
  type = "horizontal",
  series = "anxiety"
)

plot(visibility, "series")
```

The source-series view restores magnitude and temporal order. It is a
diagnostic complement to the network rather than an alternative network model.

## Weighted distance network

The visibility graph above is unweighted: an edge is either present or absent.
`tsn()` with `method = "distance"` instead joins sliding windows of the series
and assigns each edge a weight equal to `1 / (1 + distance)`, so the same
`"network"` plot type now carries continuous structure. The `connect = "nearest"`
rule retains only each window's two nearest neighbours, which keeps the figure
legible.

```{r distance-network, eval=has_cograph, fig.alt="Nearest-neighbour distance network over sliding windows of the short anxiety series; node size encodes degree and edges join the most similar windows."}
windows <- tsn(
  data = short,
  method = "distance",
  series = "anxiety",
  step = 2,
  connect = "nearest",
  neighbors = 2
)

plot(windows)
```

Node size still encodes degree, but the edges now express graded similarity
rather than presence alone. The same plot type therefore reports a different
kind of structure once the underlying model changes from visibility to distance:
the constructor, not the plot argument, determines what an edge means.

## Adjusting the network encoding

Graphical arguments change how a network is drawn without changing the fitted
object. Here the visibility graph is redrawn with a circular layout; the nodes,
edges, and weights are identical to the default view above.

```{r network-encoding, eval=has_cograph, fig.alt="The horizontal visibility graph redrawn with a circular layout, showing that the layout rearranges the same twenty-five nodes and forty-one edges."}
visibility <- vg(
  data = short,
  type = "horizontal",
  series = "anxiety"
)

plot(visibility, layout = "circle")
```

The circular layout places the observations evenly around a ring, but the graph
itself is unchanged. This separation matters for interpretation: a claim about
central or peripheral observations must rest on the edges stored in the model,
not on the visual prominence a particular layout happens to give a node.

# Transition Network Analysis plots

The TNA family comprises models returned by `ts_tna()`, `ts_ftna()`,
`ts_cna()`, and `ts_atna()`. Each model supports the same three plot types.

| `type` | Graphical encoding | Scientific use |
|---|---|---|
| `"combined"` | Source sequence and transition network in one figure | Connecting temporal state membership to estimated state relationships |
| `"network"` | Transition network only | Examining edge weights, direction, and persistence |
| `"series"` | State-classified source sequence only | Inspecting the sequence used to estimate transition relationships |

The `type` argument selects the overall representation. Within the combined
view, `network_width` controls the relative width of the network panel.
`overlay` and `ribbon` determine how states are displayed with the source
series, while `points`, `alpha`, and `palette` control its appearance. Network
encoding is modified with `node_size`, `node_scale`, and `show_weights`.
Arguments not consumed by the tsn method are passed to `cograph::splot()`.

The `network` argument selects a per-series or summary network. These coincide
for the single source series used here. The distinction becomes relevant only
when a model contains more than one source sequence.

## Combined

The state sequence is converted to a transition network and immediately
plotted. The combined representation is the default.

```{r tna-combined, eval=has_nestimate && has_cograph, fig.width=9}
probabilities <- ts_tna(states)

plot(probabilities)
```

The temporal panel preserves the observations and their assigned states. The
network panel represents states as nodes and conditional transition
probabilities as directed edges. Their juxtaposition makes the derivation of
the relational structure explicit.

## Network

```{r tna-network, eval=has_nestimate && has_cograph}
probabilities <- ts_tna(states)

plot(probabilities, "network")
```

The network-only view allocates the full plotting region to the relational
structure. Directed edges encode conditional transition probabilities and
self-loops encode persistence.

## Source series

```{r tna-series, eval=has_nestimate, fig.width=8}
probabilities <- ts_tna(states)

plot(probabilities, "series")
```

The series-only view retains the state classification but suppresses the
network. It is used to inspect the sequence from which transition weights were
estimated.

# Extracted-model plots

`series_networks()` creates an indexed collection of source-specific network
models. Its `series` argument selects a model when the collection contains more
than one source. With the single series used here, no selection argument is
required. The model is extracted and plotted in the same code block.

```{r extracted-network, eval=has_nestimate && has_cograph}
probabilities <- ts_tna(states)
individual <- series_networks(probabilities)

plot(individual)
```

The plot uses the TNA network representation and retains the provenance stored
by `series_networks()`.

# Selecting a plot

The plot type should follow the scientific question rather than the desired
appearance alone.

| Scientific question | Recommended view |
|---|---|
| When do discrete states or local directions occur? | State overlay, ribbon, or trend points |
| How was a classification produced? | Trend panels or a stacked state view |
| What temporal sequence generated the network? | Source-series view |
| What relationships define the constructed network? | Network view |
| How does a state sequence relate to its transition structure? | Combined TNA view |

Arguments controlling colour, size, labels, and layout change the graphical
encoding but do not change the fitted object. Consequently, interpretive claims
should follow the construction method and edge definition stored in the model,
not the visual prominence created by a particular style option.
