---
title: "One Series, Every Lens: Network Analysis of 50 Days of Steps"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{One Series, Every Lens: Network Analysis of 50 Days of Steps}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 7, fig.height = 4.2, dpi = 150, out.width = "100%"
)
has_nestimate <- requireNamespace("Nestimate", quietly = TRUE)
has_cograph <- requireNamespace("cograph", quietly = TRUE)
```

A time series hides more structure than its line plot shows. The same 50
numbers contain *levels* (which days were low, which high), *direction*
(when the person was improving or declining), *geometry* (which peaks
dominate their surroundings), *similarity* (which stretches of days
resemble each other), and *dynamics* (how likely a low day is to follow
a high one). Each of these questions has a network representation, and
tsn builds all of them from one input with one family of verbs.

This vignette runs every analysis type in the package on a single short
series — the first 50 days of one participant's daily step counts — so
you can see what each lens adds that the others miss:

1. **Levels** — `discretize()` turns values into ordered states and
   plots them as shading, ribbons, and guide lines.
2. **Direction** — `trend()` classifies each day as ascending,
   descending, flat, or turbulent from a rolling slope.
3. **Geometry** — `vg()` builds visibility graphs whose hubs are the
   days that dominate the series' skyline.
4. **Similarity** — `tsn(unit = "window")` connects stretches of days
   that behave alike, revealing regimes.
5. **Dynamics** — `ts_tna()` builds a state transition network and
   opens the door to compatible Nestimate analyses.

| Question | Verb | Returns |
|---|---|---|
| Which days were low/mid/high? | `discretize()` | `tsn_states` (tidy states) |
| When was the person improving? | `trend()` | `tsn_trend` (tidy classification) |
| Which days dominate the skyline? | `vg()` | `tsn` network (tidy dyads) |
| Which stretches resemble each other? | `tsn(unit = "window")` | `tsn` network (tidy dyads) |
| How do states follow each other? | `ts_tna()` | Nestimate `netobject` |

# The series

The bundled `steps` data holds daily step counts for 151 individuals.
One participant's first 50 complete days are enough to demonstrate
everything:

```{r data}
library(tsn)
data(steps)
steps_50 <- head(subset(steps, id == 536), 50)
head(steps_50, 3)
```

# Levels: from values to states

Step counts on their own are hard to compare across days — 11,000 steps
is meaningful only relative to what this person usually does.
`discretize()` makes the comparison explicit by cutting the series into
ordered states. Quantile cuts target similarly occupied states, although ties
at a cut point can make their counts unequal:

```{r discretize}
states <- discretize(steps_50$steps, method = "quantile", n_states = 3,
                     labels = c("low", "mid", "high"))
summary(states)
```

The default plot shades the series by state, which answers a question
the line plot cannot: not *what* the values were, but *when the person
occupied each regime and for how long*. Adding `lines = "horizontal"`
draws the cut points, so you can also see how close each day came to
switching state:

```{r states-overlay}
plot(states, series = "series_1", lines = "horizontal")
```

The ribbon view moves the same information into a strip under a clean
line, which is the form you want when the values themselves matter and
the shading would cover them:

```{r states-ribbon}
plot(states, "ribbon", points = TRUE)
```

# Direction: trend classification

States say where the series *is*; they say nothing about where it is
*going* — a mid day during recovery and a mid day during decline are
the same state but opposite situations. `trend()` separates them by
classifying each day from a rolling regression slope:

```{r trend}
classified <- trend(steps_50$steps, window = 7)
summary(classified)
```

For this participant the 50 days split into
`r sum(as.character(classified$state) == "Ascending")` ascending and
`r sum(as.character(classified$state) == "Descending")`
descending days with a turbulent stretch — direction changes are
frequent, which already predicts what the transition network will show
at the end of this vignette: little persistence outside the high state.
The panel view stacks the values above the rolling slope they were
classified on, so every label is traceable to the metric that produced
it:

```{r trend-panels, fig.height = 5}
plot(classified, "panels")
```

Because states and trend answer complementary questions, the most
informative single picture of the series combines them: one clean line
with both classifications as ribbons underneath. Disagreements between
the two strips are the interesting days — high-state days that are
already descending, low-state days that are ascending:

```{r multi-ribbon, fig.height = 4.6}
plot(states, "ribbon", series = "series_1",
     ribbons = list(Trend = classified), points = TRUE)
```

# Geometry: visibility graphs

A visibility graph connects two days if a straight line between their
values clears every day in between — the series' skyline turned into a
network. Its value is that peaks become *hubs*: a single exceptional
day is visible from many others, so degree concentrates on the days
that dominate their surroundings, and the degree distribution
distinguishes periodic, random, and fractal series:

```{r nvg}
nvg <- vg(steps_50$steps)
summary(nvg)
```

The network object contains a tidy dyad table, available directly through
`as.data.frame()`:

```{r nvg-edges}
head(as.data.frame(nvg), 3)
```

The horizontal variant (`vg(x, "horizontal")`) only connects days that
see each other over a *flat* line. Its edges are a subset of the natural
visibility edges (and can coincide with them for some series), it has known
theoretical baselines (an i.i.d. series has asymptotic mean degree 4), and it
is often useful when comparing ordering structure across series:

```{r hvg}
vg(steps_50$steps, "horizontal")
```

With cograph installed, any tsn network renders directly through
`cograph::splot()` — tsn results are dual-class cograph objects:

```{r nvg-splot, eval = has_cograph, fig.height = 5}
cograph::splot(nvg, layout = "spring", labels = FALSE,
               scale_nodes_by = "degree", node_size_range = c(2, 9),
               node_fill = "#4E79A7", edge_label_style = "none",
               edge_color = "#B9C2CC")
```

Node size is degree, so the large nodes *are* the skyline-dominating
days — the hubs the geometry lens contributes. Visibility also
composes with discretization: `tsn(x, "quantile")` collapses the
visibility network onto states, connecting states whose days see each
other:

```{r state-visibility}
tsn(steps_50$steps, "quantile")
```

# Similarity: windowed distance networks

The lenses so far treat single days. To ask which *stretches* of days
resemble each other, `tsn()` slides a window along the series and
connects windows by similarity. With a 10-day window stepping by 5,
the 50 days become 9 overlapping ten-day stretches; keeping the closest
quarter of pairs leaves only the strongest resemblances:

```{r window}
win <- tsn(steps_50$steps, method = "distance", unit = "window",
           window = 10, step = 5, distance = "correlation",
           connect = "percentile", percentile = 0.25)
summary(win)
```

Edges between non-adjacent windows are the finding here: they mean the
participant *returned* to an earlier pattern — regime recurrence that
no chronological plot shows. The `from_start`/`from_end` columns map
every window back to its days:

```{r window-edges}
head(as.data.frame(win), 3)
```

Setting `chain = TRUE` asks a different question of the same windows:
not "which windows are alike" but "how much does each ten-day stretch differ
from the next". The chain of consecutive distances is a change profile
of the series — spikes in `distance` locate the moments the routine
shifted. Dynamic time warping tolerates a few days' misalignment, so
the spikes reflect genuine pattern change rather than shifted timing:

```{r chain}
chain <- tsn(steps_50$steps, method = "distance", unit = "window",
             window = 10, step = 5, distance = "dtw", chain = TRUE)
as.data.frame(chain)
```

Thirteen other distance measures are available in addition to correlation and
DTW, including cross-correlation (`"ccf"`) and information-theoretic
(`"nmi"`, `"voi"`) distances. The event-based measures (`"event_sync"` and
`"van_rossum"`) instead expect each series unit to be a strictly increasing
vector of event times. The choice of `distance` determines what "similar"
means; its input must still satisfy that measure's domain.

# Dynamics: state transition networks

The state sequence from the levels section is itself a process: does
this person *stay* low once they are low? `ts_tna()` discretizes the
series and hands the state sequence to Nestimate, returning a full
transition-network model in which edges are the probabilities of
moving between states:

```{r ts-tna, eval = has_nestimate}
network <- ts_tna(steps_50$steps, n_states = 3,
                  labels = c("low", "mid", "high"))
round(network$weights, 2)
```

The rows answer the persistence question directly: high days repeat
(`high → high` is
`r if (has_nestimate) sprintf("%.2f", network$weights["high", "high"]) else "0.62"`),
while low and mid days are nearly coin flips among destinations — the
volatility the trend section anticipated. The default plot places the
shaded series beside its transition network, sharing one colour
mapping so the shading, the strip, and the network nodes read as one
story. Nodes are sized by in-strength (self-loops excluded), so the
state the person keeps *arriving in* is the visually dominant one:

```{r ts-tna-plot, eval = has_nestimate && has_cograph, fig.width = 10, fig.height = 4}
plot(network, ribbon = TRUE)
```

When probabilities are not enough, `ts_ftna()` exposes the raw transition
counts, making the denominator behind each row of probabilities visible:

```{r ts-ftna, eval = has_nestimate}
counts <- ts_ftna(steps_50$steps, n_states = 3,
                  labels = c("low", "mid", "high"))
counts$weights
```

Because the result is a genuine Nestimate `netobject` that keeps its source
data, compatible Nestimate analyses apply without conversion. Centralities
summarize each state's role in the dynamics — in-strength ranks states by how
much of the traffic flows into them:

```{r centrality, eval = has_nestimate}
Nestimate::net_centrality(network)
```

One confirmatory tool deliberately does *not* apply yet:
`Nestimate::bootstrap_network()` resamples whole sequences, so a single
series gives it nothing to resample — Nestimate warns and returns
degenerate intervals. Validation becomes possible exactly when the next
section's step is taken.

# From one series to many

Every verb above accepts many series as readily as one — a named list,
a wide matrix, or long data with `id`/`time`/`value` columns. With
several series, the transition-network plot pairs each series with its
**own** network built from its own transitions (`series_networks()`
splits the pooled model formally, sharing one node set so the networks
are directly comparable), and `network = "summary"` pools everything
into a single model instead:

```{r two-series, eval = has_nestimate && has_cograph, fig.width = 10, fig.height = 6}
two <- rbind(head(subset(steps, id == 536), 50),
             head(subset(steps, id == 88), 50))
pair <- ts_tna(two, value = "steps", id = "id", time = "day",
               n_states = 3, labels = c("low", "mid", "high"))
plot(pair, ribbon = TRUE, overlay = "none")
```

With sequences to resample, the bootstrap now runs — and its verdict is
itself the finding:

```{r bootstrap, eval = has_nestimate}
boot <- Nestimate::bootstrap_network(pair, iter = 200, seed = 1)
boot
```

No edge is certified stable, because two 50-day series are thin
evidence for nine transition probabilities. That is the bootstrap doing
its job: the pooled matrix above is a description of these 100 days,
not yet a validated model of the behavior — add participants and the
intervals tighten.

Two participants, two different dynamics, one shared colour and node
vocabulary — which is the point of the whole design: every lens in this
vignette used the same compact family of verbs, and every result provides
tidy tabular access and plotting methods. The 50 days never changed; only the
question did.
